---
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# mismeasured

High-performance SIMEX and MC-SIMEX correction for measurement error and misclassification in generalized linear models.

## Installation

```r
# install.packages("remotes")
remotes::install_github("OJALAB/mismeasured")
```

## Overview

The `mismeasured` package implements the Simulation-Extrapolation (SIMEX) and Misclassification SIMEX (MC-SIMEX) algorithms for correcting bias due to measurement error or misclassification in regression models. It provides:

- A **unified `simex()` function** with brms-style formula terms: `me()` for continuous measurement error and `mc()` for discrete misclassification
- **C++ simulation engine** via Rcpp/RcppEigen (100-300x faster than pure-R implementations)
- **Standard and improved MC-SIMEX** (Sevilimedu & Yu, 2026) with exact correction factors
- Standard S3 methods: `summary()`, `plot()`, `predict()`, `confint()`, `refit()`, etc.

## Quick start

### Continuous measurement error

When a covariate is measured with additive Gaussian error, wrap it with `me(variable, sd)`:

```{r simex-me}
library(mismeasured)

set.seed(42)
n <- 2000
x_true <- rnorm(n)
y <- 1 + 2 * x_true + rnorm(n, sd = 0.5)
x_obs <- x_true + rnorm(n, sd = 0.5)  # observed with error
df <- data.frame(y = y, x = x_obs)

fit <- simex(y ~ me(x, 0.5), data = df, B = 200)
summary(fit)
```

### Misclassification

When a discrete covariate is subject to misclassification, wrap it with `mc(variable, matrix)`:

```{r simex-mc}
z_true <- rbinom(n, 1, 0.4)
y2 <- rpois(n, exp(0.5 + 0.8 * z_true + 0.3 * x_true))

# Misclassify z
z_star <- z_true
z_star[z_true == 0] <- rbinom(sum(z_true == 0), 1, 0.10)
z_star[z_true == 1] <- 1 - rbinom(sum(z_true == 1), 1, 0.15)

Pi <- matrix(c(0.9, 0.1, 0.15, 0.85), 2, 2)
df2 <- data.frame(y = y2, z = factor(z_star), x = x_true)

# Improved MC-SIMEX (default) -- only needs B=1
fit_mc <- simex(y ~ mc(z, Pi) + x, family = poisson(), data = df2)
summary(fit_mc)
```

Use `method = "standard"` for the classic extrapolation-based MC-SIMEX:

```{r simex-mc-std}
fit_std <- simex(y ~ mc(z, Pi) + x, family = poisson(), data = df2,
                 method = "standard", B = 200)
summary(fit_std)
```

## Formula syntax

| Term | Meaning | Example |
|------|---------|---------|
| `me(x, 0.5)` | `x` measured with Gaussian error, sd = 0.5 | `y ~ me(x, 0.5) + w` |
| `me(x, sd_x)` | Heteroscedastic error, sd from column `sd_x` | `y ~ me(x, sd_x) + w` |
| `mc(z, Pi)` | `z` is a misclassified factor, Pi is the K x K misclassification matrix | `y ~ mc(z, Pi) + x` |

## References

- Cook, J.R. and Stefanski, L.A. (1994). Simulation-extrapolation estimation in parametric measurement error models. *JASA*, 89, 1314--1328.
- Kuchenhoff, H., Mwalili, S.M. and Lesaffre, E. (2006). A general method for dealing with misclassification in regression: The misclassification SIMEX. *Biometrics*, 62(1), 85--96.
- Sevilimedu, V. and Yu, L. (2026). An improved misclassification simulation extrapolation (MC-SIMEX) algorithm. *Statistics in Medicine*, 45, e70418.
