---
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# mismeasured

<!-- badges: start -->
[![R-CMD-check](https://github.com/OJALAB/mismeasured/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/OJALAB/mismeasured/actions/workflows/R-CMD-check.yaml)
[![Codecov](https://codecov.io/gh/OJALAB/mismeasured/graph/badge.svg)](https://codecov.io/gh/OJALAB/mismeasured)
<!-- badges: end -->

Bias correction for generalized linear models with measurement error and misclassification.

## Installation

```r
# install.packages("remotes")
remotes::install_github("OJALAB/mismeasured")
```

## Overview

The `mismeasured` package provides two complementary approaches for correcting bias in GLMs when covariates are measured with error or subject to misclassification:

**`simex()`** --- Simulation-Extrapolation (SIMEX / MC-SIMEX) with a formula interface:

- `me()` terms for continuous measurement error, `mc()` terms for discrete misclassification
- C++ simulation engine via Rcpp/RcppEigen (100--300x faster than pure-R implementations)
- Standard and improved MC-SIMEX (Sevilimedu & Yu, 2026) with exact fixed-matrix correction

**`mcglm()`** --- Analytical bias correction for GLMs with misclassified covariates (Battaglia, Christensen, Hansen & Sacher, 2025):

- **Naive** --- uncorrected GLM on the proxy covariate
- **BCA** --- additive bias correction
- **BCM** --- multiplicative bias correction (iterated BCM converges to CS)
- **CS** --- corrected-score estimator
- **One-step** --- joint mixture-likelihood via automatic differentiation (RTMB)
- Supports binary and multicategory misclassified covariates, Poisson/Binomial/Gaussian families, and multinomial response models
- Asymptotic inference (sandwich SE, Wald CIs) and the usual glm-style
  S3 methods (`summary`, `vcov`, `confint`, `fitted`, `predict`,
  `residuals`, `logLik`, `AIC`, ...) are provided for every method

## Quick start

### Continuous measurement error (SIMEX)

When a covariate is measured with additive Gaussian error, wrap it with `me(variable, sd)`:

```{r simex-me}
library(mismeasured)

set.seed(42)
n <- 2000
x_true <- rnorm(n)
y <- 1 + 2 * x_true + rnorm(n, sd = 0.5)
x_obs <- x_true + rnorm(n, sd = 0.5)  # observed with error
df <- data.frame(y = y, x = x_obs)

fit <- simex(y ~ me(x, 0.5), data = df, B = 200)
summary(fit)
```

### Misclassification (MC-SIMEX)

When a discrete covariate is subject to misclassification, wrap it with `mc(variable, matrix)`:

```{r simex-mc}
z_true <- rbinom(n, 1, 0.4)
y2 <- rpois(n, exp(0.5 + 0.8 * z_true + 0.3 * x_true))

# Misclassify z
z_star <- z_true
z_star[z_true == 0] <- rbinom(sum(z_true == 0), 1, 0.10)
z_star[z_true == 1] <- 1 - rbinom(sum(z_true == 1), 1, 0.15)

Pi <- matrix(c(0.9, 0.1, 0.15, 0.85), 2, 2)
df2 <- data.frame(y = y2, z = factor(z_star), x = x_true)

# Improved MC-SIMEX (default) -- only needs B=1
fit_mc <- simex(y ~ mc(z, Pi) + x, family = poisson(), data = df2)
summary(fit_mc)
```

### Bias-corrected GLM (mcglm)

For analytical bias corrections when misclassification rates are known:

```{r mcglm-example}
set.seed(42)
n <- 5000
z_true <- rbinom(n, 1, 0.4)
x1 <- rnorm(n)
y <- rpois(n, exp(0.8 * z_true - 0.5 + 0.7 * x1))

# Introduce misclassification
p01 <- 0.10; p10 <- 0.15
z_hat <- z_true
z_hat[z_true == 0] <- rbinom(sum(z_true == 0), 1, p01)
z_hat[z_true == 1] <- 1 - rbinom(sum(z_true == 1), 1, p10)

Pi <- matrix(c(1 - p01, p01, p10, 1 - p10), 2, 2)
df3 <- data.frame(y = y, z = factor(z_hat), x1 = x1)

fit <- mcglm(y ~ mc(z, Pi) + x1, data = df3, family = "poisson",
             method = c("naive", "bca", "bcm", "cs"), pi_z = 0.4)
fit
```

#### Inference and glm-style methods

`mcglm()` returns asymptotic standard errors for every fitted method
(sandwich estimators from Battaglia et al., 2025). All the usual GLM
S3 methods are available; pass `method =` to select an estimator.

```{r mcglm-inference}
# Wald table per method (estimate, SE, z, p)
summary(fit)

# Variance-covariance matrix and confidence intervals for the CS estimator
vcov(fit, method = "cs")
confint(fit, method = "cs", level = 0.95)

# Standard glm helpers, dispatched per-method
coef(fit, method = "bca")
head(fitted(fit, method = "cs"))
head(residuals(fit, method = "cs", type = "pearson"))
AIC(fit)         # naive log-likelihood when no onestep was fit
nobs(fit); family(fit)$family
```

## Formula syntax (simex)

| Term | Meaning | Example |
|------|---------|---------|
| `me(x, 0.5)` | `x` measured with Gaussian error, sd = 0.5 | `y ~ me(x, 0.5) + w` |
| `me(x, sd_x)` | Heteroscedastic error, sd from column `sd_x` | `y ~ me(x, sd_x) + w` |
| `mc(z, Pi)` | `z` is a misclassified factor, Pi is the K x K misclassification matrix | `y ~ mc(z, Pi) + x` |

## References

- Battaglia, L., Christensen, T., Hansen, S. and Sacher, S. (2025). Inference for regression with variables generated by AI or machine learning. *arXiv preprint arXiv:2402.15585*.
- Cook, J.R. and Stefanski, L.A. (1994). Simulation-extrapolation estimation in parametric measurement error models. *JASA*, 89, 1314--1328.
- Kuechenhoff, H., Mwalili, S.M. and Lesaffre, E. (2006). A general method for dealing with misclassification in regression: The misclassification SIMEX. *Biometrics*, 62(1), 85--96.
- Sevilimedu, V. and Yu, L. (2026). An improved misclassification simulation extrapolation (MC-SIMEX) algorithm. *Statistics in Medicine*, 45, e70418.
