---
title: "Bias-corrected GLMs with misclassified covariates: mcglm() and mclm()"
author: "Maciej Beresewicz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bias-corrected GLMs with misclassified covariates: mcglm() and mclm()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6, fig.height = 4
)
library(mismeasured)
set.seed(42)
f3 <- function(x) formatC(x, format = "f", digits = 3)
f4 <- function(x) formatC(x, format = "f", digits = 4)
```

## Scope

This vignette describes the analytical estimators of `mcglm()` (and
`mclm()`, its Gaussian wrapper) for GLMs in which a discrete covariate is
observed only through a misclassified proxy: the naive fit, the additive
and multiplicative bias corrections (BCA, BCM), the drift-corrected score
(CS), the subtraction correction (SUB, the default), the expectation
correction (EC), the induced likelihood (IL), the corrected score of
Akazawa, Kinukawa and Nakamura (CS-AKN) and a one-step mixture likelihood.

Here the misclassification probabilities are taken as known. When they are
estimated from an audit, see `vignette("validation", "mismeasured")`; for
many categories and small audits, `vignette("sparse", "mismeasured")`; for
simulation-extrapolation, `vignette("simex", "mismeasured")`.

## Setup and notation

Let $Y_i$ be a scalar response, $x_i \in \mathbb{R}^r$ a vector of cleanly
observed covariates (typically including an intercept), and
$Z_i \in \{0, 1, \dots, K-1\}$ a latent categorical regressor observed only
through a proxy $\hat Z_i$. For $K = 2$ take
$\xi_i = (Z_i, x_i^\top)^\top$ and $\hat\xi_i = (\hat Z_i, x_i^\top)^\top$;
for $K \ge 3$, $Z_i$ is dummy-encoded with baseline $0$. The parameter is
$\psi = (\gamma^\top, \alpha^\top)^\top$, and
$\mathbb{E}[Y_i \mid Z_i, x_i] = \mu(\psi^\top \xi_i)$ with $\mu$ the
inverse canonical link.

**Misclassification.** $\Pi_{j\ell} = \Pr(\hat Z = j \mid Z = \ell)$ (for
$K = 2$: $p_{01} = \Pi_{21}$, $p_{10} = \Pi_{12}$), prevalence
$\pi_\ell = \Pr(Z = \ell)$, and predictive probabilities
$W_{j\ell} = \Pr(Z = \ell \mid \hat Z = j) =
\Pi_{j\ell}\pi_\ell / \sum_m \Pi_{jm}\pi_m$. Misclassification is
nondifferential, $\hat Z_i \perp\!\!\!\perp (Y_i, x_i) \mid Z_i$.

**Prevalence.** All methods except CS-AKN use one prevalence $\pi$ for
every observation, i.e. they assume that $Z$ is independent of $x$. When
the latent category depends on the covariates they are biased unless a
covariate-dependent prevalence $\Pr(Z \mid x)$ is modelled
(`control_mc(prevalence_model = ~ ...)`, see the validation vignette).

## The naive estimator and its drift

The naive estimator solves the proxy score
$$
\hat U_n(\psi) = \frac{1}{n} \sum_{i=1}^n \hat\xi_i \{Y_i - \mu(\psi^\top \hat\xi_i)\} = 0 ,
$$
i.e. a GLM with $\hat Z_i$ in place of $Z_i$. For binary $Z$ its expectation
at the truth is $\mathbb{E}[m_i(\psi_0)]$, with the per-observation drift
$$
m_i(\psi) = \begin{pmatrix} -c_1\, \delta_i(\psi) \\ -c_2\, \delta_i(\psi)\, x_i \end{pmatrix},
\qquad
\delta_i(\psi) = \mu(\gamma + \alpha^\top x_i) - \mu(\alpha^\top x_i),
$$
$c_1 = p_{01}(1-\pi)$ and $c_2 = p_{01}(1-\pi) - p_{10}\pi$ (Warmuz and
Beresewicz, 2026; the multicategory drift uses $(\Pi, \pi)$ in the same
way). The drift is the conditional expectation of the naive score given the
covariates, $m_i(\psi) = \mathbb{E}_\psi[\hat\xi_i\{Y_i - \mu(\psi^\top\hat\xi_i)\} \mid x_i]$.

## The estimators

| Method | Estimating function or idea | Needs | Assumes $Z \perp x$ |
|---|---|---|---|
| `naive` | proxy GLM | -- | -- (inconsistent) |
| `bca` | $\hat\psi - \hat I^{-1}\hat m(\hat\psi)$ | $\Pi, \pi$ (or $c_1, c_2$) | yes |
| `bcm` | $\hat\psi - (\hat I + \hat M)^{-1}\hat m(\hat\psi)$ | $\Pi, \pi$ (or $c_1, c_2$) | yes |
| `cs` | $\sum_i [\hat\xi_i\{Y_i - \mu(\psi^\top\hat\xi_i)\} - m_i(\psi)] = 0$ | $\Pi, \pi$ (or $c_1, c_2$) | yes |
| `sub` | $\sum_i \hat\xi_i\{Y_i - \nu_i(\psi)\} = 0$, $\nu_i = \sum_\ell W_{\hat z_i\ell}\,\mu(\gamma_\ell + \alpha^\top x_i)$ | $\Pi, \pi$ | yes (or a prevalence model) |
| `ec` | $\sum_i \sum_\ell \omega_{i\ell}(\psi) S_{i\ell}(\psi) = 0$, $\omega_{i\ell} \propto W_{\hat z_i\ell} f(Y_i \mid \ell, x_i)$ | $\Pi, \pi$, response distribution | yes (or a prevalence model) |
| `il` | maximise $\prod_i \sum_\ell W_{\hat z_i\ell} f(Y_i \mid \ell, x_i)$ | $\Pi, \pi$, response distribution | yes (or a prevalence model) |
| `cs_akn` | corrected score: $\sum_i \sum_\ell \tilde x_{i\ell} S_{i\ell}(\psi) = 0$, $\tilde x_{i\cdot}$ the $\hat z_i$-th column of $\Pi^{-1}$ | $\Pi$ only | no |
| `onestep` | maximise $\prod_i \sum_\ell \omega_{\hat z_i\ell} f(Y_i \mid \ell, x_i)$ with free or fixed weights | -- (`fix_omega = TRUE`: $\Pi, \pi$) | yes |

Here $S_{i\ell}(\psi)$ is the complete-data score at $Z_i = \ell$ and
$f(\cdot \mid \ell, x)$ the response density (Poisson, Bernoulli, or normal
with an estimated $\sigma$). `ec`, `il` and `onestep` need a count response
for Poisson and a 0/1 response for the binomial family.

### How they relate

* **`sub` dominates `cs`.** `cs` subtracts the conditional expectation of
  the naive score given $x$; `sub` (Yi, Yan, Liao and Spiegelman, 2019,
  Section 3.2) subtracts it given $(\hat Z, x)$. The two estimating
  functions differ by a term with conditional mean zero given $x$, so they
  have the same Jacobian and the middle matrix of `cs` is that of `sub`
  plus the variance of that term: `sub` is never less efficient. Despite
  its name, `cs` is not a corrected score in the sense of Nakamura (1990);
  the label is kept for compatibility.
* **`ec` and `il` coincide with known probabilities.** The EC estimating
  function, the posterior mean of the complete-data score (Section 3.3), is
  the score of the induced likelihood (Section 3.1; their Appendix A), and
  both equal `onestep` with `fix_omega = TRUE`. They use the full response
  distribution and are the efficient estimators when it is correctly
  specified. They differ once the probabilities are estimated: `ec` plugs
  them in, `il` estimates them jointly with $\psi$.
* **Structural versus functional.** `cs`, `sub`, `ec` and `il` are
  structural: they need the distribution of $Z$ given $x$ (here a constant
  prevalence, or a prevalence model). `cs_akn` (Akazawa, Kinukawa and
  Nakamura, 1998; Yi et al., 2019, eq. 17) is functional: its conditional
  expectation given $(Y, Z, x)$ is the complete-data score whatever the
  distribution of $Z$, so it needs only $\Pi$ and stays consistent when $Z$
  depends on $x$. The price is a larger variance and an inversion of
  $Q = [p_1 - p_0, \dots, p_{K-1} - p_0]$ that becomes unstable when the
  classifier is weak (Yi et al., 2019, Section 3.5).
* **`bca` and `bcm`** are one-step approximations to `cs` derived under
  misclassification probabilities that shrink with $n$ (Battaglia et al.,
  2025); with fixed misclassification `bcm` is close to `cs` and `bca`
  keeps a first-order bias.

### Inference

With known probabilities:

* naive, BCA, BCM: $A^{-1} C A^{-1}$ (drifting-regime formula);
  `vcov_corrected = TRUE` gives a more conservative BCA/BCM sandwich;
* CS, SUB, EC, CS-AKN: the Z-estimator sandwich $J^{-1} S J^{-\top}$ (EC
  with the Louis-identity Jacobian);
* onestep: inverse Hessian of the integrated log-likelihood.

With a validation sample, every corrected method propagates the
uncertainty of the estimated probabilities through stacked estimating
functions (or the joint likelihood for `il`), and bootstrap or posterior
variances are available; see the validation vignette.

## Worked examples

### Binary misclassification, Poisson response

```{r poisson-binary}
n  <- 5000
Pi <- matrix(c(0.9, 0.1, 0.1, 0.9), 2, 2)   # p01 = p10 = 0.10
z  <- rbinom(n, 1, 0.4)
z_hat <- z
z_hat[z == 0] <- rbinom(sum(z == 0), 1, 0.10)
z_hat[z == 1] <- 1 - rbinom(sum(z == 1), 1, 0.10)
x1 <- rnorm(n)
y  <- rpois(n, exp(0.8 * z - 0.5 + 0.7 * x1))

df <- data.frame(y = y, z = factor(z_hat), x1 = x1)

methods <- c("naive", "bca", "bcm", "cs", "sub", "ec", "il", "cs_akn")
fit <- mcglm(y ~ mc(z, Pi) + x1, data = df, family = "poisson",
             method = methods, pi_z = 0.4)
gamma_hat <- sapply(methods, function(m) coef(fit, method = m)[["gamma"]])
se_gamma  <- sapply(methods, function(m) fit$se[[m]][["gamma"]])
round(rbind(estimate = gamma_hat, se = se_gamma), 3)
```

```{r poisson-numbers, include = FALSE}
fixed <- c("cs", "sub", "ec", "il", "cs_akn")
```

The true effect is $\gamma_0 = 0.8$. The naive estimate,
`r f3(gamma_hat[["naive"]])`, is attenuated; the corrected estimates lie
between `r f3(min(gamma_hat[-1]))` and `r f3(max(gamma_hat[-1]))`. Among the
estimators valid under fixed misclassification, the standard error of
`sub` (`r f4(se_gamma[["sub"]])`) is no larger than that of `cs`
(`r f4(se_gamma[["cs"]])`); `ec` and `il` coincide
(`r f3(gamma_hat[["ec"]])`) and, using the Poisson distribution of the
response, have the smallest standard error
(`r f4(min(se_gamma[fixed]))`); `cs_akn` has
`r f4(se_gamma[["cs_akn"]])`. The standard errors of `bca` and `bcm` come
from the drifting-misclassification theory and are not comparable.

```{r poisson-summary}
coef(summary(fit))$sub
```

### Binary misclassification, logistic response

```{r logistic}
n  <- 5000
Pi <- matrix(c(0.85, 0.15, 0.10, 0.90), 2, 2)
z  <- rbinom(n, 1, 0.4); zh <- z
zh[z == 0] <- rbinom(sum(z == 0), 1, 0.15)
zh[z == 1] <- 1 - rbinom(sum(z == 1), 1, 0.10)
x1 <- rnorm(n)
y  <- rbinom(n, 1, plogis(0.9 * z - 0.4 + 0.6 * x1))

dfb <- data.frame(y = y, z = factor(zh), x1 = x1)
fitb <- mcglm(y ~ mc(z, Pi) + x1, data = dfb, family = "binomial",
              method = c("naive", "sub", "il", "cs_akn"), pi_z = 0.4)
round(do.call(rbind, fitb$coefficients), 3)
```

### Multicategory misclassification ($K = 3$)

For $K \ge 3$ pass the misclassification matrix and the prevalence vector;
`cs_akn` needs only $\Pi$.

```{r multicat}
n   <- 5000
Pi3 <- matrix(0.1, 3, 3); diag(Pi3) <- 0.8
pi3 <- c(0.5, 0.3, 0.2)
Z  <- sample(0:2, n, replace = TRUE, prob = pi3)
Zh <- vapply(Z, function(z) sample(0:2, 1, prob = Pi3[, z + 1]), 0L)
x1 <- rnorm(n)
ym <- rpois(n, exp(c(0, 1.0, -0.9)[Z + 1] + 0.8 - 0.7 * x1))

fit3 <- mcglm(ym, z_hat = Zh, x = cbind(1, x1), family = "poisson",
              method = c("naive", "sub", "il", "cs_akn"),
              Pi = Pi3, pi_z = pi3)
round(do.call(rbind, fit3$coefficients), 3)
```

The true effects are $\gamma = (1.0, -0.9)$.

### Confidence intervals

```{r ci}
confint(fit, method = "sub")
confint(fit, method = "cs_akn")
```

### One-step (joint mixture likelihood)

With `fix_omega = FALSE` (default) the one-step estimator estimates the
mixture weights jointly with $\psi$ from the main study alone, so it needs
no misclassification probabilities but relies on the mixture being
identified by the response distribution; with `fix_omega = TRUE` and
known probabilities it equals `il` and `ec`. It uses `RTMB`.

```{r onestep, eval = requireNamespace("RTMB", quietly = TRUE)}
fit_os <- mcglm(y ~ mc(z, Pi) + x1, data = df, family = "poisson",
                method = c("naive", "onestep"))
coef(fit_os, method = "onestep")
```

## `mclm()`: Gaussian-fixed convenience wrapper

`mclm()` is a thin wrapper around `mcglm()` with `family = "gaussian"`, in
the spirit of base R's `lm()` versus `glm()`. All other arguments forward
unchanged and the returned object remains of class `"mcglm"`.

```{r mclm}
n  <- 4000
Pi <- matrix(c(0.9, 0.1, 0.1, 0.9), 2, 2)
z  <- rbinom(n, 1, 0.4); zh <- z
zh[z == 0] <- rbinom(sum(z == 0), 1, 0.10)
zh[z == 1] <- 1 - rbinom(sum(z == 1), 1, 0.10)
x1 <- rnorm(n)
yg <- 0.8 * z - 0.5 + 0.7 * x1 + rnorm(n, 0, 0.5)

dfg <- data.frame(y = yg, z = factor(zh), x1 = x1)
fitg <- mclm(y ~ mc(z, Pi) + x1, data = dfg,
             method = c("naive", "sub", "il", "cs_akn"), pi_z = 0.4)
round(do.call(rbind, fitg$coefficients), 3)
```

For the Gaussian family `sub` and `cs_akn` have closed forms (linear
estimating equations), and `il` estimates the residual standard deviation
jointly (`fitg$nuisance$il$sigma`).

## Choosing a method

* **Default: `sub`.** It needs only the mean model, is never less efficient
  than `cs`, and supports every validation design.
* **Response distribution trusted: `il` or `ec`.** They are the efficient
  estimators for a correctly specified Poisson, Bernoulli or normal
  response. With an estimated misclassification matrix `il` also lets the
  main study inform it, and for an internal audit it needs no design
  weights when selection depends only on observed variables.
* **$Z$ may depend on the covariates:** add a prevalence model to
  `sub`/`ec`/`il` (`control_mc(prevalence_model = ~ ...)`), or use
  `cs_akn`, which needs only $\Pi$. Without either, all other methods are
  biased; with an internal audit `mcglm()` tests for this association and
  warns.
* **Only $\Pi$ is credible (e.g. an external benchmark with another
  prevalence):** `cs_akn`, or the other methods with the prevalence
  re-estimated from the main study (`control_mc(prevalence = "em")`).
* **Small, shrinking misclassification:** `bca`/`bcm` are valid in that
  regime; with fixed misclassification `bca` keeps a first-order bias.
* **No information on the misclassification:** `onestep` with free mixture
  weights, if the response distribution identifies the mixture.
* **Multinomial response:** `naive` and `onestep` only.

## Result object and S3 methods

`mcglm()` and `mclm()` return objects of class `"mcglm"`:

- `print()`, `summary()` -- coefficient tables per method;
- `coef()`, `vcov()`, `confint()` and `se.mcglm()` -- with `method = ` to
  select the estimator; `fit$se` holds all standard errors;
- `predict()`, `fitted()`, `residuals()` -- on the original (proxy) data;
- `logLik()`, `AIC()`, `BIC()`, `nobs()` -- for the naive and one-step fits;
- `fit$nuisance`, `fit$mc_estimate`, `fit$convergence` -- estimated
  misclassification quantities and solver status.

## References

Akazawa, K., Kinukawa, N. and Nakamura, T. (1998). A note on the corrected
score function adjusting for misclassification. *Journal of the Japan
Statistical Society*, 28(1), 115--123. doi:10.14490/jjss1995.28.115.

Battaglia, L., Christensen, T., Hansen, S. and Sacher, S. (2025). Inference
for regression with variables generated by AI or machine learning. *arXiv
preprint* arXiv:2402.15585. doi:10.48550/arXiv.2402.15585.

Nakamura, T. (1990). Corrected score function for errors-in-variables
models: methodology and application to generalized linear models.
*Biometrika*, 77(1), 127--137. doi:10.1093/biomet/77.1.127.

Warmuz, M. and Beresewicz, M. (2026). Bias corrections for GLMs with
misclassified covariates. Working paper.

Yi, G. Y., Yan, Y., Liao, X. and Spiegelman, D. (2019). Parametric
regression analysis with covariate misclassification in main
study/validation study designs. *The International Journal of
Biostatistics*, 15(1), 20170002. doi:10.1515/ijb-2017-0002.
