---
title: "Bias-corrected GLMs with misclassified covariates: mcglm() and mclm()"
author: "Maciej Beresewicz, Michal Warmuz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bias-corrected GLMs with misclassified covariates: mcglm() and mclm()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6, fig.height = 4
)
library(mismeasured)
set.seed(42)
```

## Scope

This vignette describes the analytical bias-correction estimators in
**mismeasured**: `mcglm()` for any supported family and `mclm()`, the
Gaussian-fixed wrapper of `mcglm()`. Five estimators are implemented for
GLMs in which a discrete covariate is observed only through a misclassified
proxy: naive, additive bias correction (BCA), multiplicative bias correction
(BCM), corrected score (CS, two formulations), and a one-step joint
mixture-likelihood estimator.

For simulation-extrapolation methods, see the companion vignette
`vignette("simex", "mismeasured")`.

## Setup and notation

Let $Y_i$ be a scalar response, $x_i \in \mathbb{R}^r$ a vector of cleanly
observed covariates (typically including an intercept), and
$Z_i \in \{0, 1, \dots, K-1\}$ a latent categorical regressor observed only
through a proxy $\hat Z_i$. For binary problems ($K = 2$) take
$\xi_i = (Z_i, x_i^\top)^\top$ and
$\hat\xi_i = (\hat Z_i, x_i^\top)^\top$; for $K \ge 3$, $Z_i$ is dummy-encoded
with baseline $0$ so $\xi_i = (d_i^\top, x_i^\top)^\top$. The parameter is
$\psi = (\gamma^\top, \alpha^\top)^\top$ with $\gamma \in \mathbb{R}^{K-1}$ on
the (dummy-coded) latent regressor and $\alpha \in \mathbb{R}^r$ on $x_i$.

The conditional mean is $\mathbb{E}[Y_i \mid Z_i, x_i] = \mu(\psi^\top \xi_i)$
for $\mu = g^{-1}$ the inverse canonical link.

**Misclassification mechanism.** For $K = 2$ this is summarized by
$p_{01} = \Pr(\hat Z = 1 \mid Z = 0)$, $p_{10} = \Pr(\hat Z = 0 \mid Z = 1)$,
and prevalence $\pi = \Pr(Z = 1)$. For $K \ge 3$, by the $K \times K$
column-stochastic matrix $\Pi_{j\ell} = \Pr(\hat Z = j \mid Z = \ell)$ and
the prevalence vector $(\pi_0, \dots, \pi_{K-1})$. We assume nondifferential
misclassification: $\hat Z_i \perp\!\!\!\perp (Y_i, x_i) \mid Z_i$.

## The naive proxy-score estimator and its bias

The naive estimator solves the proxy-score equation
$$
\hat U_n(\psi) = \frac{1}{n} \sum_{i=1}^n \hat\xi_i \{Y_i - \mu(\psi^\top \hat\xi_i)\} = 0.
$$
This is just GLM estimation with $\hat Z_i$ in place of $Z_i$. It is
typically inconsistent for $\psi_0$. For binary $Z$, the package theory
(Warmuz & Beresewicz, 2026, Proposition on score drift) shows
$$
\mathbb{E}[\hat U_n(\psi_0)] = \mathbb{E}[m_i(\psi_0)],
$$
where the per-observation drift is
$$
m_i(\psi) = \begin{pmatrix} -c_1\, \delta_i(\psi) \\ -c_2\, \delta_i(\psi)\, x_i \end{pmatrix},
\qquad
\delta_i(\psi) = \mu(\gamma + \alpha^\top x_i) - \mu(\alpha^\top x_i),
$$
with constants $c_1 = p_{01}(1-\pi)$ and $c_2 = p_{01}(1-\pi) - p_{10}\pi$.
The multicategory version replaces $(\delta_i, c_1, c_2)$ with explicit
double sums over $(\Pi, \pi_z)$ (Appendix B of the package theory note).

## The five estimators

| Method      | Idea                                                                                              | Required nuisance | Asymptotic regime                |
|-------------|---------------------------------------------------------------------------------------------------|-------------------|----------------------------------|
| `naive`     | proxy GLM, no correction                                                                          | --                | inconsistent under fixed $\Pi$   |
| `bca`       | additive: $\hat\psi_{\mathrm{bca}} = \hat\psi - \hat I(\hat\psi)^{-1}\hat m(\hat\psi)$              | $(c_1, c_2)$ or $(\Pi, \pi)$ | drifting $\Pi = O(n^{-1/2})$ |
| `bcm`       | multiplicative: $\hat\psi_{\mathrm{bcm}} = \hat\psi - (\hat I + \hat M)^{-1} \hat m$                | same              | drifting $\Pi = O(n^{-1/2})$ |
| `cs`        | corrected score: solve $\sum_i [\hat\xi_i\{Y_i - \mu(\psi^\top\hat\xi_i)\} - m_i(\psi)] = 0$       | same              | **fixed $\Pi$**                  |
| `cs_akn`    | AKN98 corrected score via unbiased surrogate $x = Q^{-1}(u - p_0)$                                | $\Pi$ only        | **fixed $\Pi$**                  |
| `onestep`   | maximize $\prod_i \sum_\ell \pi_\ell \Pi_{\hat z_i+1, \ell+1} f(Y_i \mid Z = \ell, \psi)$ via RTMB | (see below)        | fixed $\Pi$                      |

`bca` and `bcm` remove the leading $n^{-1/2}$ bias under a drifting-sequence
regime where misclassification probabilities shrink with $n$. `cs` and
`cs_akn` are unbiased estimating equations under fixed $\Pi$ -- they remain
consistent even when misclassification is moderate or large. `onestep` is a
likelihood-based alternative implemented through automatic differentiation
(`RTMB`); it estimates mixture weights jointly with $\psi$ unless
`fix_omega = TRUE`.

### Two CS formulations

The package implements *two* corrected-score estimators that solve the same
population equation but differ in finite-sample form:

- **`cs` (drift)** keeps the proxy as-is and adds the explicit drift
  $m_i(\psi)$ to the naive score. Needs $(\Pi, \pi_z)$. (Warmuz &
  Beresewicz, 2026, Theorem on the corrected-score estimator;
  Battaglia et al., 2025 for the linear-model origin.)
- **`cs_akn` (unbiased surrogate)** transforms the regressor via
  $x = Q^{-1}(u - p_0)$ where
  $Q = [p_1 - p_0, \ldots, p_{K-1} - p_0]$, then substitutes into the
  log-likelihood. Needs only $\Pi$ -- $\pi_z$ cancels out.
  (Akazawa, Kinukawa & Nakamura, 1998.)

The two are population-equivalent at $\psi_0$. Empirically they agree to
$\sim 10^{-4}$ for $K = 2$ and within sampling noise for $K \ge 3$;
`cs_akn`'s SD is slightly larger when $\Pi$ has off-diagonals away from zero
(the $Q^{-1}$ amplifies single-observation contributions).

### Inference

All five estimators report an asymptotic variance through `vcov()`:

- naive / BCA / BCM: $A^{-1} C A^{-1}$ where $A = \mathbb{E}[\dot\mu(\psi_0^\top\xi_i)\xi_i\xi_i^\top]$ and $C = \mathbb{E}[\varepsilon_i^2 \xi_i \xi_i^\top]$.
- CS / cs_akn: $J^{-1} S J^{-\top}$ with $J = -(\hat I + \hat M)$ (cs) or $J = -I_{\mathrm{akn}}$ (cs_akn) and $S = \mathbb{E}[\phi_i \phi_i^\top]$.
- onestep: inverse Hessian of the integrated log-likelihood.

Set `vcov_corrected = TRUE` in `mcglm()` to get a more conservative BCA/BCM
sandwich that accounts for the additional uncertainty from estimating
$\hat m(\hat\psi)$.

## Worked examples

### Binary misclassification, Poisson response

```{r poisson-binary}
n  <- 5000
Pi <- matrix(c(0.9, 0.1, 0.1, 0.9), 2, 2)   # p01 = p10 = 0.10
z  <- rbinom(n, 1, 0.4)
z_hat <- z
z_hat[z == 0] <- rbinom(sum(z == 0), 1, 0.10)
z_hat[z == 1] <- 1 - rbinom(sum(z == 1), 1, 0.10)
x1 <- rnorm(n)
y  <- rpois(n, exp(0.8 * z - 0.5 + 0.7 * x1))

df <- data.frame(y = y, z = factor(z_hat), x1 = x1)

fit <- mcglm(y ~ mc(z, Pi) + x1, data = df, family = "poisson",
             method = c("naive", "bca", "bcm", "cs", "cs_akn"))
fit
```

The naive $\hat\gamma$ is biased toward zero; all four corrections recover
$\gamma_0 = 0.8$.

```{r poisson-summary}
summary(fit)
```

### Binary misclassification, logistic response

```{r logistic}
n  <- 5000
Pi <- matrix(c(0.85, 0.15, 0.10, 0.90), 2, 2)
z  <- rbinom(n, 1, 0.4); zh <- z
zh[z == 0] <- rbinom(sum(z == 0), 1, 0.15)
zh[z == 1] <- 1 - rbinom(sum(z == 1), 1, 0.10)
x1 <- rnorm(n)
y  <- rbinom(n, 1, plogis(0.9 * z - 0.4 + 0.6 * x1))

dfb <- data.frame(y = y, z = factor(zh), x1 = x1)
fitb <- mcglm(y ~ mc(z, Pi) + x1, data = dfb, family = "binomial",
              method = c("naive", "cs", "cs_akn"))
coef(fitb, method = "naive")
coef(fitb, method = "cs")
coef(fitb, method = "cs_akn")
```

### Multicategory misclassification ($K = 3$)

For $K \ge 3$ pass the misclassification matrix and (for `cs`/`bca`/`bcm`)
the prevalence vector $\pi$. `cs_akn` needs only $\Pi$.

```{r multicat}
n   <- 5000
Pi3 <- matrix(0.1, 3, 3); diag(Pi3) <- 0.8
pi3 <- c(0.5, 0.3, 0.2)
Z  <- sample(0:2, n, replace = TRUE, prob = pi3)
Zh <- vapply(Z, function(z) sample(0:2, 1, prob = Pi3[, z + 1]), 0L)
x1 <- rnorm(n)
ym <- rpois(n, exp(c(0, 1.0, -0.9)[Z + 1] + 0.8 - 0.7 * x1))

fit3 <- mcglm(ym, z_hat = Zh, x = cbind(1, x1), family = "poisson",
              method = c("naive", "cs", "cs_akn"),
              Pi = Pi3, pi_z = pi3)
coef(fit3, method = "cs")
coef(fit3, method = "cs_akn")
```

### Confidence intervals

```{r ci}
confint(fit, method = "cs")
confint(fit, method = "cs_akn")
```

### One-step (joint mixture likelihood)

The one-step estimator maximizes the integrated mixture likelihood across
the latent categories. It uses `RTMB` for automatic differentiation and is
robust under fixed $\Pi$ for any $K$. Mixture weights can be estimated
jointly (default) or fixed at user-supplied $\pi_z$ via `fix_omega = TRUE`.

```{r onestep, eval = requireNamespace("RTMB", quietly = TRUE)}
fit_os <- mcglm(y ~ mc(z, Pi) + x1, data = df, family = "poisson",
                method = c("naive", "onestep"))
coef(fit_os, method = "onestep")
```

## `mclm()`: Gaussian-fixed convenience wrapper

`mclm()` is a thin wrapper around `mcglm()` with `family = "gaussian"`, in
the spirit of base R's `lm()` versus `glm()`. All other arguments forward
unchanged and the returned object remains of class `"mcglm"`, so all S3
methods (`summary`, `vcov`, `confint`, ...) work identically.

```{r mclm}
n  <- 4000
Pi <- matrix(c(0.9, 0.1, 0.1, 0.9), 2, 2)
z  <- rbinom(n, 1, 0.4); zh <- z
zh[z == 0] <- rbinom(sum(z == 0), 1, 0.10)
zh[z == 1] <- 1 - rbinom(sum(z == 1), 1, 0.10)
x1 <- rnorm(n)
yg <- 0.8 * z - 0.5 + 0.7 * x1 + rnorm(n, 0, 0.5)

dfg <- data.frame(y = yg, z = factor(zh), x1 = x1)
fitg <- mclm(y ~ mc(z, Pi) + x1, data = dfg,
             method = c("naive", "cs", "cs_akn"))
coef(fitg, method = "cs")
coef(fitg, method = "cs_akn")
```

For Gaussian, `cs_akn` admits a closed-form linear-system solution (no
iterative solver), so the two CS formulations agree to numerical precision.

## When to use which method

- **Reliable $(\Pi, \pi_z)$, well-curated proxy.** Any of `cs`, `cs_akn`, or
  `onestep`. Use `cs` for the most established formulation; `onestep` when
  you also want a likelihood-based confidence set or model fit comparison.
- **Reliable $\Pi$ but $\pi_z$ unknown or unstable.** Prefer `cs_akn` --
  the surrogate construction makes $\pi_z$ irrelevant.
- **Misclassification probabilities small and shrinking with $n$.** `bca`
  or `bcm` are valid (drifting-sequence regime). `bcm` removes second-order
  bias and is closer to the iterated CS update.
- **Multicategory $K \ge 3$.** All five estimators apply; `cs_akn` carries a
  small variance penalty (≈ 10--20% larger SD) versus drift `cs`.
- **Multinomial response.** Only `naive` and `onestep` (the others are
  univariate-response specific).

## Estimated misclassification parameters

The package theory (Warmuz & Beresewicz, 2026, Proposition on plug-in
robustness) shows that the corrected-score estimator remains consistent
when $(\Pi, \pi)$ are estimated from auxiliary data, provided the plug-in
score satisfies $\sup_\psi \|\Phi_n^* - \Phi_n^\circ\| = o_P(1)$, and is
asymptotically unbiased under the tighter rate $o_P(n^{-1/2})$. In
practice this means a validation subsample of moderate size (e.g. 5--10%
of the main sample) is sufficient.

## Result object and S3 methods

`mcglm()` and `mclm()` both return objects of class `"mcglm"`. The
following S3 methods are available:

- `print()`, `summary()` -- coefficient tables per method.
- `coef()`, `vcov()`, `se()`, `confint()` -- accept a `method = ` argument
  to select which estimator's quantity to extract.
- `predict()`, `fitted()`, `residuals()` -- on the original (proxy) data.
- `logLik()`, `AIC()`, `BIC()`, `nobs()` -- only meaningful for `onestep`.

## References

Akazawa, K., Kinukawa, N. and Nakamura, T. (1998). A note on the corrected
score function adjusting for misclassification. *J. Japan Statist. Soc.*
28(1), 115--123. (Foundational existence theorem and explicit corrected
scores for misclassified discrete covariates; basis of `cs_akn`.)

Battaglia, L., Christensen, T., Hansen, S. and Sacher, S. (2025). Inference
for regression with variables generated by AI or machine learning.
*arXiv preprint* arXiv:2402.15585. (Drifting-asymptotics framework for
linear models; origin of BCA/BCM and one-step.)

Nakamura, T. (1990). Corrected score function for errors-in-variables
models: methodology and application to generalized linear models.
*Biometrika* 77(1), 127--137. \doi{10.1093/biomet/77.1.127} (Originating
reference for the corrected-score concept under continuous additive error.)

Warmuz, M. and Beresewicz, M. (2026). Bias corrections for GLMs with
misclassified covariates. Working paper. (GLM extension of the
drifting-asymptotics regime; multicategory drift formulas; plug-in
robustness conditions used in the package.)
