---
title: "Getting started with mismeasured"
author: "Maciej Beresewicz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with mismeasured}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6, fig.height = 4
)
library(mismeasured)
set.seed(42)
```

## What this package does

**mismeasured** corrects regression estimates for measurement error and
misclassification of covariates (and response). It bundles two complementary
toolsets:

| Toolset                     | When to use                                                                         | Vignette                          |
|-----------------------------|--------------------------------------------------------------------------------------|-----------------------------------|
| `simex()`                   | Continuous measurement error, Berkson error, single or multiple misclassified factors, response misclassification | `vignette("simex", "mismeasured")`  |
| `mcglm()` / `mclm()`        | Misclassified discrete covariate(s) with known (or auxiliary-estimated) $\Pi$, where you want analytical bias-corrected GLM/LM inference | `vignette("mcglm", "mismeasured")`  |

Variables subject to error are marked directly in the formula:
`me(x, sd)` for continuous measurement error, `mc(z, Pi)` for discrete
misclassification.

### Supported response families

`gaussian()`, `poisson()`, `binomial()` (canonical links). `mcglm()` also
supports `family = "multinomial"` for the naive and one-step estimators.

## A SIMEX example in 30 seconds

```{r simex-quickstart}
n <- 1500
x_true <- rnorm(n)
y      <- 1 + 2 * x_true + rnorm(n, sd = 0.5)
x_obs  <- x_true + rnorm(n, sd = 0.5)
df     <- data.frame(y = y, x = x_obs)

fit_simex <- simex(y ~ me(x, 0.5), data = df, B = 200, seed = 42)

c(naive_lm = coef(lm(y ~ x, data = df))["x"],
  simex    = coef(fit_simex)["x"],
  truth    = 2)
```

The naive OLS slope is attenuated; SIMEX recovers a value close to the
true slope of 2. See `vignette("simex", "mismeasured")` for the full
algorithm, MC-SIMEX, response misclassification, Berkson error, and weights.

## A bias-corrected GLM example in 30 seconds

```{r mcglm-quickstart}
n  <- 3000
Pi <- matrix(c(0.9, 0.1, 0.1, 0.9), 2, 2)
z      <- rbinom(n, 1, 0.4)
z_hat  <- z
z_hat[z == 0] <- rbinom(sum(z == 0), 1, 0.10)
z_hat[z == 1] <- 1 - rbinom(sum(z == 1), 1, 0.10)
x1 <- rnorm(n)
y  <- rpois(n, exp(0.8 * z - 0.5 + 0.7 * x1))
df <- data.frame(y = y, z = factor(z_hat), x1 = x1)

fit_mcglm <- mcglm(y ~ mc(z, Pi) + x1, data = df, family = "poisson",
                   method = c("naive", "cs", "cs_akn"))
coef(fit_mcglm, method = "naive")
coef(fit_mcglm, method = "cs")
coef(fit_mcglm, method = "cs_akn")
```

The naive $\hat\gamma$ is attenuated below the truth (0.8); both corrected
scores recover it. `mclm()` is a Gaussian-fixed wrapper of `mcglm()` for
linear models. See `vignette("mcglm", "mismeasured")` for the full theory,
the BCA/BCM/one-step alternatives, and the multicategory case.

## Utility functions

The package also exports two helpers for working with misclassification
matrices, used by both toolsets:

- `build.mc.matrix()` finds the nearest valid (non-negative-power) matrix
  given an empirical estimate of $\Pi$.
- `check.mc.matrix()` validates a list of candidate $\Pi$s.

```{r utils}
Pi_empirical <- matrix(c(0.85, 0.10, 0.05,
                         0.10, 0.80, 0.10,
                         0.05, 0.10, 0.85), 3, 3, byrow = TRUE)
check.mc.matrix(list(Pi_empirical))
Pi_valid <- build.mc.matrix(Pi_empirical)
check.mc.matrix(list(Pi_valid))
```

## Getting help

- `?simex`, `?mcglm`, `?mclm`, `?me`, `?mc` — function references.
- `vignette("simex", "mismeasured")` — SIMEX/MC-SIMEX in depth.
- `vignette("mcglm", "mismeasured")` — bias-corrected GLM theory and methods.
- `https://ojalab.github.io/mismeasured/` — pkgdown site with the full
  reference index.
- Issues: `https://github.com/OJALAB/mismeasured/issues`.
