---
title: "Correcting measurement error and misclassification with mismeasured"
author: "Maciej Beresewicz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Correcting measurement error and misclassification with mismeasured}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6, fig.height = 4
)
library(mismeasured)
set.seed(42)
```

## Introduction

The **mismeasured** package provides Simulation-Extrapolation (SIMEX) and
Misclassification SIMEX (MC-SIMEX) corrections for generalized linear models
where covariates are measured with error or subject to misclassification.

The package uses a formula interface inspired by **brms**: error-prone variables
are marked directly in the formula with `me()` (continuous measurement error) or
`mc()` (discrete misclassification). The simulation step runs in C++ via Rcpp and
RcppEigen for high performance.

### Supported error types

| Type | Formula syntax | Method |
|------|---------------|--------|
| Continuous measurement error | `me(x, sd)` | SIMEX (Cook & Stefanski, 1994) |
| Berkson measurement error | `me(x, sd, type = "berkson")` | Berkson SIMEX |
| Systematic (non-zero mean) ME | `me(x, sd, mean = mu)` | Non-zero mean SIMEX |
| Discrete misclassification | `mc(z, Pi)` | MC-SIMEX (Kuchenhoff et al., 2006) |
| Improved misclassification | `mc(z, Pi)` with `method = "improved"` | Sevilimedu & Yu (2026) |
| Response misclassification | `mc(y, Pi_y) ~ ...` | MC-SIMEX on response |

### Supported families

Gaussian (`gaussian()`), Poisson (`poisson()`), and Binomial (`binomial()`).

## Background: the SIMEX algorithm

When a covariate $X$ is measured with additive error $W = X + U$, where
$U \sim N(0, \sigma_u^2)$, regression coefficients are biased toward zero
(attenuation bias). SIMEX corrects this in two steps:

1. **Simulation**: For $\lambda \in \{0.5, 1, 1.5, 2\}$ and $b = 1, \ldots, B$,
   generate $W^*_{b,\lambda} = W + \sqrt{\lambda} \cdot U_b$, where
   $U_b \sim N(0, \sigma_u^2)$. Refit the model on each pseudo-dataset.

2. **Extrapolation**: Fit a polynomial (quadratic, linear, or loglinear) to the
   average estimates $\bar{\hat\beta}(\lambda)$ as a function of $\lambda$, and
   extrapolate to $\lambda = -1$ (no measurement error).

For misclassification, the MC-SIMEX algorithm replaces additive noise with
resampling from $\Pi^\lambda$, the misclassification matrix raised to fractional
powers.

## Example 1: Continuous measurement error

```{r me-example}
n <- 2000
x_true <- rnorm(n)
y <- 1 + 2 * x_true + rnorm(n, sd = 0.5)
x_obs <- x_true + rnorm(n, sd = 0.5)  # measurement error with sd = 0.5
df <- data.frame(y = y, x = x_obs)

# Naive OLS (biased)
naive <- lm(y ~ x, data = df)

# SIMEX corrected
fit <- simex(y ~ me(x, 0.5), data = df, B = 200, seed = 42)

cat("True slope:  2.000\n")
cat("Naive slope:", round(coef(naive)["x"], 3), "\n")
cat("SIMEX slope:", round(coef(fit)["x"], 3), "\n")
```

The naive estimate is attenuated toward zero. SIMEX recovers a value closer to
the true slope of 2.

### Summary and inference

```{r me-summary}
summary(fit)
confint(fit)
```

### SIMEX trajectory plot

```{r me-plot}
plot(fit)
```

The plot shows coefficient estimates at each $\lambda$ level (solid points),
the naive estimate at $\lambda = 0$ (open circle), the extrapolation curve,
and the corrected estimate at $\lambda = -1$ (red cross).

### Changing the extrapolation method

```{r me-refit}
fit_lin <- refit(fit, extrapolation = "linear")
cat("Quadratic:", round(coef(fit)["x"], 3), "\n")
cat("Linear:   ", round(coef(fit_lin)["x"], 3), "\n")
```

## Example 2: Multiple error-prone covariates

```{r multi-me}
x1_true <- rnorm(n)
x2_true <- rnorm(n)
y <- 1 + 1.5 * x1_true - 0.8 * x2_true + rnorm(n, sd = 0.5)
df2 <- data.frame(
  y  = y,
  x1 = x1_true + rnorm(n, sd = 0.3),
  x2 = x2_true + rnorm(n, sd = 0.5)
)

fit2 <- simex(y ~ me(x1, 0.3) + me(x2, 0.5), data = df2, B = 200, seed = 42)
summary(fit2)
```

## Example 3: Misclassification (standard MC-SIMEX)

A binary covariate $Z$ is misclassified with known misclassification matrix
$\Pi$, where $\Pi_{jl} = P(Z^* = j \mid Z = l)$.

```{r mc-standard}
z <- rbinom(n, 1, 0.4)
y3 <- 1 + 2 * z + 0.5 * rnorm(n) + rnorm(n)

# Misclassify: sensitivity = 0.85, specificity = 0.9
z_star <- z
z_star[z == 0] <- rbinom(sum(z == 0), 1, 0.10)
z_star[z == 1] <- 1 - rbinom(sum(z == 1), 1, 0.15)

Pi <- matrix(c(0.9, 0.1, 0.15, 0.85), 2, 2)
df3 <- data.frame(y = y3, z = factor(z_star))

fit3 <- simex(y ~ mc(z, Pi), data = df3, method = "standard",
              B = 200, seed = 42)

cat("True z effect:     2.000\n")
cat("Naive z effect:   ", round(fit3$naive.coefficients[1], 3), "\n")
cat("MC-SIMEX z effect:", round(coef(fit3)[1], 3), "\n")
```

## Example 4: Improved MC-SIMEX (exact correction)

The improved method (Sevilimedu & Yu, 2026) computes an exact correction factor
instead of extrapolation. It requires only B = 1 replicate and is the default
for single binary misclassified covariates.

```{r mc-improved}
fit4 <- simex(y ~ mc(z, Pi), data = df3, seed = 42)  # default: method = "improved"

cat("Method:", fit4$method, "\n")
cat("Correction factor c(lambda):", round(fit4$c.lambda, 4), "\n")
cat("Improved z effect:", round(coef(fit4)[1], 3), "\n")
```

### Optimal lambda

The improved method can search for the lambda that minimizes the absolute
correction factor:

```{r mc-optimal}
fit4_opt <- simex(y ~ mc(z, Pi), data = df3, lambda = "optimal", seed = 42)
cat("Optimal lambda:", fit4_opt$lambda[-1], "\n")
cat("Optimal c(lambda):", round(fit4_opt$c.lambda, 4), "\n")
```

## Example 5: Multiple misclassified covariates

```{r multi-mc}
z1 <- rbinom(n, 1, 0.4)
z2 <- rbinom(n, 1, 0.6)
y5 <- 1 + 2 * z1 + 1.5 * z2 + rnorm(n)

# Misclassify both
z1_star <- z1
z1_star[z1 == 0] <- rbinom(sum(z1 == 0), 1, 0.10)
z1_star[z1 == 1] <- 1 - rbinom(sum(z1 == 1), 1, 0.15)

z2_star <- z2
z2_star[z2 == 0] <- rbinom(sum(z2 == 0), 1, 0.08)
z2_star[z2 == 1] <- 1 - rbinom(sum(z2 == 1), 1, 0.12)

Pi1 <- matrix(c(0.9, 0.1, 0.15, 0.85), 2, 2)
Pi2 <- matrix(c(0.92, 0.08, 0.12, 0.88), 2, 2)
df5 <- data.frame(y = y5, z1 = factor(z1_star), z2 = factor(z2_star))

fit5 <- simex(y ~ mc(z1, Pi1) + mc(z2, Pi2), data = df5,
              method = "standard", B = 200, seed = 42)
summary(fit5)
```

With multiple `mc()` terms, the standard method is used automatically. Each
variable is resampled independently via its own misclassification matrix.

## Example 6: Response misclassification

When the outcome itself is misclassified (e.g., disease status from an imperfect
diagnostic test), use `mc()` on the left-hand side:

```{r response-mc}
x6 <- rnorm(n)
y6_true <- rbinom(n, 1, plogis(-0.5 + 1.2 * x6))

# Misclassify the response
y6_obs <- y6_true
y6_obs[y6_true == 0] <- rbinom(sum(y6_true == 0), 1, 0.10)
y6_obs[y6_true == 1] <- 1 - rbinom(sum(y6_true == 1), 1, 0.08)

Pi_y <- matrix(c(0.90, 0.10, 0.08, 0.92), 2, 2)
df6 <- data.frame(y = factor(y6_obs), x = x6)

fit6 <- simex(mc(y, Pi_y) ~ x, family = binomial(), data = df6,
              B = 200, seed = 42)

cat("True x effect:     1.200\n")
cat("Naive x effect:   ", round(fit6$naive.coefficients["x"], 3), "\n")
cat("MC-SIMEX x effect:", round(coef(fit6)["x"], 3), "\n")
```

Response misclassification can be combined with covariate misclassification:

```{r response-plus-covariate}
z6 <- rbinom(n, 1, 0.4)
y6b_true <- rbinom(n, 1, plogis(-0.5 + 1.0 * z6 + 0.5 * x6))
y6b_obs <- y6b_true
y6b_obs[y6b_true == 0] <- rbinom(sum(y6b_true == 0), 1, 0.10)
y6b_obs[y6b_true == 1] <- 1 - rbinom(sum(y6b_true == 1), 1, 0.08)

z6_star <- z6
z6_star[z6 == 0] <- rbinom(sum(z6 == 0), 1, 0.10)
z6_star[z6 == 1] <- 1 - rbinom(sum(z6 == 1), 1, 0.15)

df6b <- data.frame(y = factor(y6b_obs), z = factor(z6_star), x = x6)

fit6b <- simex(mc(y, Pi_y) ~ mc(z, Pi) + x, family = binomial(),
               data = df6b, B = 200, seed = 42)
summary(fit6b)
```

## Example 7: Berkson measurement error

In the Berkson error model, the true value is a noisy version of the observed
(assigned) value: $X = W + U$. This arises in occupational exposure assessment,
where $W$ is the assigned group mean and $X$ is the individual's true exposure.

```{r berkson}
w <- rnorm(n)
u <- rnorm(n, sd = 0.5)
x_true_b <- w + u
y7 <- 1 + 2 * x_true_b + rnorm(n, sd = 0.5)
df7 <- data.frame(y = y7, x = w)

fit7_class <- simex(y ~ me(x, 0.5), data = df7, B = 200, seed = 42)
fit7_berk  <- simex(y ~ me(x, 0.5, type = "berkson"), data = df7,
                    B = 200, seed = 42)

cat("Classical SIMEX slope:", round(coef(fit7_class)["x"], 3), "\n")
cat("Berkson SIMEX slope:  ", round(coef(fit7_berk)["x"], 3), "\n")
```

## Example 8: Non-zero mean measurement error

When measurement error has a systematic component $U \sim N(\mu, \sigma^2)$
with $\mu \neq 0$:

```{r nonzero-mean}
x_true_m <- rnorm(n)
x_obs_m <- x_true_m + rnorm(n, mean = 0.3, sd = 0.5)
y8 <- 1 + 2 * x_true_m + rnorm(n, sd = 0.5)
df8 <- data.frame(y = y8, x = x_obs_m)

fit8_std <- simex(y ~ me(x, 0.5), data = df8, B = 200, seed = 42)
fit8_nzm <- simex(y ~ me(x, 0.5, mean = 0.3), data = df8, B = 200, seed = 42)

cat("Standard SIMEX (ignoring mean):", round(coef(fit8_std)["x"], 3), "\n")
cat("Non-zero mean SIMEX:           ", round(coef(fit8_nzm)["x"], 3), "\n")
```

## Example 9: Poisson regression with misclassification

```{r poisson-mc}
z9 <- rbinom(n, 1, 0.4)
y9 <- rpois(n, exp(0.5 + 0.8 * z9))

z9_star <- z9
z9_star[z9 == 0] <- rbinom(sum(z9 == 0), 1, 0.10)
z9_star[z9 == 1] <- 1 - rbinom(sum(z9 == 1), 1, 0.15)

df9 <- data.frame(y = y9, z = factor(z9_star))

fit9 <- simex(y ~ mc(z, Pi), family = poisson(), data = df9, seed = 42)
summary(fit9)
```

## Example 10: Frequency weights

The `weights` argument supports frequency (prior) weights:

```{r weights}
# Aggregated data with weights
df10_agg <- data.frame(
  y = c(2.5, 3.1, 4.2, 5.0),
  x = c(1.0, 2.0, 3.0, 4.0),
  freq = c(100, 150, 120, 130)
)

fit10 <- simex(y ~ me(x, 0.3), data = df10_agg, weights = df10_agg$freq,
               B = 100, seed = 42)
coef(fit10)
```

## Utility functions

### Building misclassification matrices

`build.mc.matrix()` finds the nearest valid misclassification matrix (one that
can be raised to fractional powers without negative probabilities):

```{r build-mc}
Pi_empirical <- matrix(c(0.85, 0.10, 0.05,
                          0.10, 0.80, 0.10,
                          0.05, 0.10, 0.85), 3, 3, byrow = TRUE)
check.mc.matrix(list(Pi_empirical))

Pi_valid <- build.mc.matrix(Pi_empirical)
check.mc.matrix(list(Pi_valid))
```

### Checking matrix validity

```{r check-mc}
P1 <- matrix(c(0.9, 0.1, 0.2, 0.8), 2, 2)
P2 <- matrix(c(0.4, 0.6, 0.6, 0.4), 2, 2)  # too much error
check.mc.matrix(list(P1, P2))
```

## Implementation details

### C++ backend

The simulation step runs entirely in C++ for all error types:

- `simex_sim_cpp`: Continuous ME with IRLS fitting + model vcov accumulation
- `mcsimex_sim_cpp`: Single-variable MC-SIMEX
- `mcsimex_multi_sim_cpp`: Multi-variable MC-SIMEX with optional response mc

The IRLS solver supports Gaussian (identity link), Poisson (log link), and
Binomial (logit link) families. Matrix exponentiation for $\Pi^\lambda$ uses
eigendecomposition.

### Variance estimation

- **ME-SIMEX**: Jackknife variance following Stefanski & Cook (1994). The
  model-based variance $\bar{V}_{\text{model}}(\lambda)$ is the average of
  $(X_{\text{sim}}'WX_{\text{sim}})^{-1}\phi$ across B refits, computed in C++.
  $V_{\text{jk}}(\lambda) = \bar{V}_{\text{model}} - \text{Cov}(\hat\theta_b)$
  is then extrapolated to $\lambda = -1$.

- **MC-SIMEX standard**: Same jackknife approach with sandwich variance at the
  mean estimate per lambda.

- **MC-SIMEX improved**: Scaled model-based variance:
  $V_{\text{corrected}} = c(\lambda)^2 \cdot V_{\text{model}}$.

## References

Cook, J.R. and Stefanski, L.A. (1994). Simulation-extrapolation estimation in
parametric measurement error models. *Journal of the American Statistical
Association*, 89(428), 1314-1328.

Kuchenhoff, H., Mwalili, S.M. and Lesaffre, E. (2006). A general method for
dealing with misclassification in regression: The misclassification SIMEX.
*Biometrics*, 62(1), 85-96.

Sevilimedu, V. and Yu, L. (2026). An improved misclassification simulation
extrapolation (MC-SIMEX) algorithm. *Statistics in Medicine*, 45, e70418.
