---
title: "Sparse validation samples: shrinkage, diagnostics and variance"
author: "Maciej Beresewicz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sparse validation samples: shrinkage, diagnostics and variance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(mismeasured)
set.seed(2027)
```

With many categories -- occupations, industries, products -- an audit of a
few hundred units leaves some true classes with a handful of observations
and many cells of the misclassification table empty. This vignette gives
the mathematics behind the estimators of `estimate_mc()` and the variance
options of `mcglm()` for that situation, and shows them on an example. For
how to describe the audit itself see `vignette("validation", "mismeasured")`.

## Notation

The latent category $Z \in \{0, \dots, K-1\}$ is observed through a proxy
$\hat Z$. The misclassification matrix, the prevalence and the predictive
matrix are
$$
\Pi_{j\ell} = \Pr(\hat Z = j \mid Z = \ell), \qquad
\pi_\ell = \Pr(Z = \ell), \qquad
W_{j\ell} = \Pr(Z = \ell \mid \hat Z = j) =
\frac{\Pi_{j\ell}\pi_\ell}{\sum_m \Pi_{jm}\pi_m}.
$$
An audit $V$ of $n_V$ units records $(Z_i, \hat Z_i)$ with design weights
$d_i$ (inverse inclusion probabilities; constant for a simple random
sample). With the weights normalised to the audit size,
$\tilde d_i = d_i\, n_V / \sum_V d_k$, the weighted counts are
$$
C_{j\ell} = \sum_{i \in V} \tilde d_i\, 1(\hat Z_i = j, Z_i = \ell), \qquad
n_\ell = \sum_j C_{j\ell}.
$$

## The column-proportion estimator and why it fails when sparse

The default (`estimator = "mle"`) is
$$
\hat\Pi_{j\ell} = C_{j\ell} / n_\ell,
\qquad \operatorname{Var}(\hat\Pi_{j\ell}) \approx
\Pi_{j\ell}(1 - \Pi_{j\ell}) / n_\ell ,
$$
the maximum-likelihood (Hajek) estimator. With small $n_\ell$:

* cells with true probability $0.01$--$0.05$ are often observed as zero;
* $\hat\Pi$ can be ill-conditioned, and every corrected estimator depends
  on it through an inverse or a ratio: `"cs_akn"` inverts
  $Q = [p_1 - p_0, \dots, p_{K-1} - p_0]$ (the rows $j \ge 1$ of $\Pi$
  minus its first column); `"sub"`, `"ec"` and `"il"` use $W$, which is
  undefined for an empty proxy row; and $\hat\pi$ obtained from the main
  study solves $\hat\Pi\pi = \hat p$.

`diagnose_mc()` reports the smallest singular value $\sigma_{\min}$ and the
condition number $\kappa(\hat\Pi) = \sigma_{\max}/\sigma_{\min}$, the
number of audited units per true class, empty cells, and $\kappa(Q)$ for
$K > 2$ or $|1 - p_{01} - p_{10}|$ for $K = 2$. The thresholds are
`control_mc(kappa_max, sigma_min, min_class_n)`; `on_ill` decides between
a warning, an error, silence, or `"regularize"`, which re-estimates with
the shrinkage estimator below.

## Empirical-Bayes shrinkage

`estimator = "eb"` gives each column of $\Pi$ a Dirichlet prior centred on
a target $T$ (a column-stochastic matrix) with concentration $c$,
$\Pi_{\cdot\ell} \sim \operatorname{Dirichlet}(c\, T_{\cdot\ell})$, and
uses the posterior mean
$$
\hat\Pi^{EB}_{j\ell} = \frac{C_{j\ell} + c\, T_{j\ell}}{n_\ell + c}
= \lambda_\ell\, \hat\Pi_{j\ell} + (1 - \lambda_\ell)\, T_{j\ell},
\qquad \lambda_\ell = \frac{n_\ell}{n_\ell + c}.
$$
Rarely audited classes ($n_\ell$ small) borrow more from the target; no
cell is zero when $T$ has no zeros.

### Targets

`target = "pooled"` uses one accuracy for all classes:
$$
T_{\ell\ell} = a = \frac{\sum_\ell C_{\ell\ell}}{\sum_{j\ell} C_{j\ell}},
\qquad T_{j\ell} = \frac{1 - a}{K - 1}\ (j \ne \ell).
$$

`target = "groups"` (with `groups = g`, a group for each category, e.g.
ISCO major groups) separates confusion within and between groups:
$$
T_{\ell\ell} = a, \qquad
T_{j\ell} = \frac{r_{\text{same}}}{|\{j' \ne \ell: g_{j'} = g_\ell\}|}\
\text{if } g_j = g_\ell, \qquad
T_{j\ell} = \frac{r_{\text{other}}}{|\{j': g_{j'} \ne g_\ell\}|}\
\text{otherwise},
$$
with $r_{\text{same}}$ and $r_{\text{other}}$ the pooled shares of
within- and between-group errors, and each column renormalised.

`target = "loglinear"` fits a Poisson quasi-independence model to the
table,
$$
\log E(C_{j\ell}) = a_j + b_\ell + \delta\, 1(j = \ell)
\;[+\ \tau\, 1(j \ne \ell, g_j = g_\ell)],
$$
and normalises the fitted values by column. `target = "matrix"` takes a
user matrix (`target_matrix`), e.g. a published confusion matrix of the
classifier.

A target close to singular would make $\hat\Pi$ unidentifiable when the
shrinkage is strong (the uniform matrix is rank one), so targets with
$\kappa(T) >$ `kappa_max` are refused.

### Concentration

With `concentration = "ml"`, $c$ maximises the Dirichlet-multinomial
marginal likelihood of the counts given $T$,
$$
\ell(c) = \sum_{\ell} \Big[ \log\Gamma(c) - \log\Gamma(n_\ell + c)
+ \sum_j \big\{ \log\Gamma(C_{j\ell} + c T_{j\ell}) - \log\Gamma(c T_{j\ell}) \big\}
\Big],
$$
over $c \in [10^{-3}, 10^{6}]$. When the target fits the data exactly the
maximum is at the upper end (full pooling); when the classes differ a lot
from the target, $c$ is small and $\hat\Pi^{EB}$ is close to the column
proportions. A fixed number can be given instead.

`estimator = "dirichlet"` adds `alpha` to every cell, i.e. $T$ uniform and
$c = K\alpha$; since the uniform matrix is singular it warns, and it is
meant only for small `alpha`.

### Inference after shrinkage

For fixed $c$ and $T$,
$$
\hat\Pi^{EB}_{j\ell} - \hat\Pi_{j\ell} =
\frac{c\,(T_{j\ell} - \hat\Pi_{j\ell})}{n_\ell + c} = O_p(n_\ell^{-1}),
$$
so the first-order (influence-function) behaviour of $\hat\Pi^{EB}$ is that
of the column proportions. `estimate_mc()` uses the estimating function
$$
\sum_{i \in V} \tilde d_i\, 1(Z_i = \ell)\{1(\hat Z_i = j) - \Pi_{j\ell}\}
+ c\,(T_{j\ell} - \Pi_{j\ell}) = 0, \qquad j \ge 1,
$$
which is solved exactly by $\hat\Pi^{EB}$. In the stacked sandwich
$A^{-1} B A^{-\top}$ the penalty $c(T - \Pi)$ enters the derivative $A$
but not the meat $B$ (it is not random), and $c$ and $T$ are treated as
fixed. The induced likelihood (`"il"`), which estimates $\Pi$ jointly with
the regression, maximises the penalised likelihood
$$
\sum_i \log \sum_\ell \Pi_{\hat z_i \ell}\,\pi_\ell\, f(y_i \mid \ell, x_i)
\;+\; \sum_{j\ell} c\, T_{j\ell} \log \Pi_{j\ell},
$$
the posterior mode under the same prior; its M-step for $\Pi$ adds
$c\,T$ to the expected counts.

## Variance options

Let $\theta = (\psi, \eta)$ stack the regression parameters $\psi$ and the
misclassification parameters $\eta$ (free entries of $\Pi$ and the
prevalence or prevalence-model coefficients).

* `variance = "delta"` (default): $\widehat{\operatorname{Var}}(\hat\theta)
  = A^{-1} B A^{-\top}$ with $A = \sum_u \partial g_u / \partial\theta^\top$
  and $B = \sum_u g_u g_u^\top$ over units $u$ (regression rows, with
  internal audit rows merged into their unit; external audit units
  separately and stratum-centred when stratified).
* `variance = "conditional"`: $\eta$ treated as known; only the $\psi$
  block of $A$ and $B$.
* `variance = "bootstrap"`: the regression rows are resampled with
  replacement -- an internal audit's units travel with their rows, which
  corresponds to Poisson sampling of the audit within the main study --
  and an external audit is resampled within its strata; $\eta$ and
  $\psi$ are re-estimated with the same settings and
  $\widehat{\operatorname{Var}}(\hat\psi) = \operatorname{Var}_b(\hat\psi^{(b)})$
  over `B` replicates.
* `variance = "posterior"` (external audits, constant prevalence): draw
  $\Pi^{(b)}_{\cdot\ell} \sim \operatorname{Dirichlet}(C_{\cdot\ell} + c\,
  T_{\cdot\ell})$ (Jeffreys $\tfrac12$ per cell without shrinkage) and
  $\pi^{(b)}$ from its Dirichlet posterior (or by EM from the main study),
  refit with $(\Pi^{(b)}, \pi^{(b)})$ treated as known, and combine
  (Rubin, 1987)
  $$
  \widehat{\operatorname{Var}}(\hat\psi) = \bar V + \Big(1 + \frac1B\Big)
  \operatorname{Var}_b(\hat\psi^{(b)}), \qquad
  \bar V = \frac1B \sum_b V_b^{\text{cond}} .
  $$
  `"il"` (which estimates $\Pi$ itself) and `"bca"`/`"bcm"` (whose
  conditional covariances understate the fixed-misclassification
  variance) keep the delta-method covariance.

The bootstrap and the posterior are useful when $n_\ell$ is so small that
the first-order approximation is doubtful; `seed` makes them reproducible
without changing the global random-number state.

## Example: six occupations and an audit of 150

Six occupations in two groups (white- and blue-collar); the classifier is
more accurate for some occupations than for others and confuses
occupations within a group three times as often as across groups.

```{r data}
K <- 6
occ <- c("clerk", "sales", "service", "craft", "operator", "elementary")
groups <- c(1, 1, 1, 2, 2, 2)
accuracy <- c(0.90, 0.82, 0.75, 0.85, 0.70, 0.88)
Pi_true <- matrix(0, K, K, dimnames = list(z_hat = occ, z = occ))
for (l in seq_len(K)) {
  w <- ifelse(groups == groups[l], 3, 1)
  w[l] <- 0
  Pi_true[, l] <- (1 - accuracy[l]) * w / sum(w)
  Pi_true[l, l] <- accuracy[l]
}
gamma_true <- c(0, 0.5, -0.3, 0.2, 0.4, -0.5)

n <- 4000
z <- sample.int(K, n, TRUE, c(0.3, 0.2, 0.15, 0.15, 0.1, 0.1)) - 1L
df <- data.frame(
  x1 = rnorm(n),
  occ = factor(occ[vapply(z, function(k)
    sample.int(K, 1, prob = Pi_true[, k + 1]), 0L)], levels = occ))
df$y <- rpois(n, exp(gamma_true[z + 1] - 0.3 + 0.5 * df$x1))
idx <- sample(n, 150)
val <- validation_sample(occ[z[idx] + 1], index = idx)
```

The column proportions:

```{r mle}
est_mle <- estimate_mc(val, z_hat = df$occ,
                       control = control_mc(on_ill = "none"))
d_mle <- diagnose_mc(est_mle)
d_mle
```

```{r mle-numbers, include = FALSE}
n_cells <- K * (K - 1)
```

`r d_mle$zero_cells` of the `r n_cells` off-diagonal cells are empty and
the rarest occupation has `r min(d_mle$n_class)` audited units. The
diagnostics report this; whether they flag a problem depends on the
conditioning of the whole matrix (`sigma_min`, `kappa_max`) and on
`min_class_n`. The empirical-Bayes estimates with the pooled and the group
targets:

```{r eb}
est_eb <- estimate_mc(val, z_hat = df$occ,
                      control = control_mc(estimator = "eb", on_ill = "none"))
est_gr <- estimate_mc(val, z_hat = df$occ,
                      control = control_mc(estimator = "eb", target = "groups",
                                           groups = groups, on_ill = "none"))
c(pooled = est_eb$shrinkage$c, groups = est_gr$shrinkage$c)
round(est_gr$Pi, 3)
err <- function(e) max(abs(unname(e$Pi) - unname(Pi_true)))
c(mle = err(est_mle), eb_pooled = err(est_eb), eb_groups = err(est_gr))
```

The regression with each estimator, using the column proportions and the
group-target shrinkage:

```{r fits}
methods <- c("sub", "ec", "il", "cs_akn")
fit_mle <- suppressWarnings(
  mcglm(y ~ mc(occ) + x1, data = df, method = methods, validation = val,
        mc_control = control_mc(on_ill = "none")))
fit_eb <- mcglm(y ~ mc(occ) + x1, data = df, method = methods,
                validation = val,
                mc_control = control_mc(estimator = "eb", target = "groups",
                                        groups = groups, on_ill = "none"))
effects <- function(fit) sapply(methods, function(m) coef(fit, method = m)[1:5])
round(cbind(true = gamma_true[-1], mle = effects(fit_mle),
            eb = effects(fit_eb)), 2)
```

A single audit says little about which estimator is more accurate; the
next section repeats the experiment.

Standard errors of the effect of `sales` under the different variance
options. The bootstrap is shown for `"sub"` only, with a small `B`, to keep
this vignette fast; it works the same way for every method but refits
`"il"` slowly.

```{r variance}
ctrl <- function(...) control_mc(estimator = "eb", target = "groups",
                                 groups = groups, on_ill = "none", ...)
fit_cond <- mcglm(y ~ mc(occ) + x1, data = df, method = c("sub", "il"),
                  validation = val, mc_control = ctrl(variance = "conditional"))
fit_boot <- mcglm(y ~ mc(occ) + x1, data = df, method = "sub",
                  validation = val,
                  mc_control = ctrl(variance = "bootstrap", B = 50, seed = 1))
se_sales <- function(fit, m) fit$se[[m]][["gamma_sales"]]
se <- rbind(delta       = c(sub = se_sales(fit_eb, "sub"), il = se_sales(fit_eb, "il")),
            conditional = c(se_sales(fit_cond, "sub"), se_sales(fit_cond, "il")),
            bootstrap   = c(se_sales(fit_boot, "sub"), NA))
round(se, 3)
```

```{r variance-numbers, include = FALSE}
f3 <- function(x) formatC(x, format = "f", digits = 3)
```

Here the three standard errors of `"sub"` are close (delta
`r f3(se["delta", "sub"])`, conditional `r f3(se["conditional", "sub"])`,
bootstrap `r f3(se["bootstrap", "sub"])`): with a concentration of
`r formatC(est_gr$shrinkage$c, format = "f", digits = 0)` the shrunk
$\hat\Pi$ draws most of its information from the group structure, so its
estimation adds little variance. The delta method treats $c$ and $T$ as
fixed; the bootstrap re-estimates them in every replicate and is the check
when the audit is this small.

## Repeated samples

The design above was repeated 100 times (new data and a new audit of 150
each time), fitting every estimator with the column proportions and with
both shrinkage targets. The code is shown but not run when the vignette is
built (it takes a few minutes):

```{r simulation, eval = FALSE}
one_run <- function(r) {
  set.seed(5000 + r)
  z <- sample.int(K, n, TRUE, c(0.3, 0.2, 0.15, 0.15, 0.1, 0.1)) - 1L
  z_hat <- vapply(z, function(k) sample.int(K, 1, prob = Pi_true[, k + 1]),
                  0L) - 1L
  x1 <- rnorm(n)
  y <- rpois(n, exp(gamma_true[z + 1] - 0.3 + 0.5 * x1))
  idx <- sample(n, 150)
  val <- validation_sample(z[idx], index = idx)
  controls <- list(
    mle = control_mc(on_ill = "none"),
    eb_pooled = control_mc(estimator = "eb", on_ill = "none"),
    eb_groups = control_mc(estimator = "eb", target = "groups",
                           groups = groups, on_ill = "none"))
  lapply(controls, function(ct) {
    fit <- suppressWarnings(mcglm(y, z_hat = z_hat, x = cbind(1, x1),
                                  method = methods, validation = val,
                                  mc_control = ct))
    c(Pi_error = max(abs(unname(fit$mc_estimate$Pi) - unname(Pi_true))),
      sapply(methods, function(m)
        sum((coef(fit, method = m)[1:5] - gamma_true[-1])^2)))
  })
}
runs <- lapply(1:100, one_run)
```

Results (SSE: sum of squared errors of the five occupation effects;
"failures": share of runs with SSE above 1, i.e. an estimate off by more
than about 0.45 on average):

| | column proportions | EB, pooled target | EB, groups target |
|---|---|---|---|
| mean of $\max_{j\ell}\lvert\hat\Pi_{j\ell} - \Pi_{j\ell}\rvert$ | 0.160 | 0.119 | 0.126 |
| median SSE, `"sub"` | 0.053 | 0.037 | 0.037 |
| median SSE, `"ec"` | 0.042 | 0.031 | 0.030 |
| median SSE, `"il"` | 0.034 | 0.028 | 0.026 |
| median SSE, `"cs_akn"` | 0.075 | 0.035 | 0.033 |
| failures, `"sub"` | 2% | 1% | 1% |
| failures, `"ec"` | 1% | 0% | 0% |
| failures, `"il"` | 0% | 0% | 0% |
| failures, `"cs_akn"` | 12% | 0% | 0% |

Shrinkage reduces the error of $\hat\Pi$ and of every estimator. The gain is
largest for `"cs_akn"`, which inverts $Q$: with the column proportions one
audit in eight gave an unusable estimate, with shrinkage none did. The
induced likelihood is the most accurate in all three columns, and the group
target, which matches how the classifier errs here, is slightly better than
the pooled one.


## Which options combine

All designs of `validation_sample()` work with all estimators of $\Pi$ and
all methods, with these exceptions (each gives an informative error):

| Option | Internal audit | External audit | Notes |
|---|---|---|---|
| `weights`, `strata` (`"z_hat"` or labels) | yes | yes (`N_strata` for strata without weights) | internal design weights cannot be combined with main-study frequency weights |
| `estimator = "ht"` | yes | yes (`N` needed) | affects only the prevalence from the audit; $\Pi$ and a prevalence model are unaffected |
| `control_mc(estimator = "eb" / "dirichlet")` | yes | yes | all targets |
| `prevalence = "validation"` | yes | yes | assumes the audit's prevalence is the main study's for external audits |
| `prevalence = "em"` | yes | yes | needs the main study's proxies (always available in `mcglm()`) |
| `prevalence = "inverse"` | yes | yes | not with `"il"` or a prevalence model |
| `prevalence_model = ~ ...` | yes | yes (`validation_sample(data = )` for `prevalence = "validation"`) | only with `"sub"`, `"ec"`, `"il"`, `"cs_akn"` |
| `variance = "delta"` / `"conditional"` | yes | yes | |
| `variance = "bootstrap"` | yes | yes | needs the `validation_sample()` itself (not an `estimate_mc()` object) and no supplied probabilities |
| `variance = "posterior"` | no | yes | constant prevalence only; `"il"`, `"bca"`, `"bcm"` keep the delta method |
| `on_ill = "regularize"` | yes | yes | |

These combinations are exercised by the package tests.

## References

Rubin, D. B. (1987). *Multiple Imputation for Nonresponse in Surveys*.
Wiley. doi:10.1002/9780470316696.

Saerens, M., Latinne, P. and Decaestecker, C. (2002). Adjusting the outputs
of a classifier to new a priori probabilities: a simple procedure.
*Neural Computation*, 14(1), 21--41. doi:10.1162/089976602753284446.

Yi, G. Y., Yan, Y., Liao, X. and Spiegelman, D. (2019). Parametric
regression analysis with covariate misclassification in main
study/validation study designs. *The International Journal of
Biostatistics*, 15(1), 20170002. doi:10.1515/ijb-2017-0002.
