---
title: "Specifying validation samples for misclassification correction"
author: "Maciej Beresewicz, Michal Warmuz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Specifying validation samples for misclassification correction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(mismeasured)
set.seed(2026)
```

## Why a validation sample

`mcglm()` corrects a regression for a covariate $Z$ that is observed only
through a proxy $\hat Z$, for example an occupation code assigned to online
job ads by a classifier. The correction needs the misclassification
probabilities

$$
\Pi_{j\ell} = \Pr(\hat Z = j \mid Z = \ell), \qquad
\pi_\ell = \Pr(Z = \ell),
$$

or equivalently the predictive probabilities
$W_{j\ell} = \Pr(Z = \ell \mid \hat Z = j)$. When they are not known, they
are estimated from a **validation (audit) sample** in which the true
category $Z$ is recorded next to $\hat Z$. How the audit was drawn decides
how it must be described to the package. This vignette walks through the
common designs.

The workflow has three steps:

1. describe the audit with `validation_sample()`;
2. optionally estimate and inspect the probabilities with `estimate_mc()`
   and `diagnose_mc()`;
3. fit the model with `mc(z, estimate)` in the formula or with
   `validation =`.

The corrected estimators of `mcglm()` use the estimated probabilities in
different ways:

| Method | Use of the validation sample |
|---|---|
| `"sub"` (default) | probabilities estimated from the audit; standard errors include their estimation uncertainty, for every design below |
| `"ec"` | as `"sub"`, with the expectation-corrected score (uses the full response distribution) |
| `"il"` | probabilities estimated *jointly* with the regression from the audit and the regression data (induced likelihood); for internal audits no design weights are needed |
| `"cs_akn"` | only $\Pi$ estimated from the audit; the prevalence is not used, so it remains valid when $Z$ depends on the covariates |
| `"cs"`, `"bca"`, `"bcm"`, `"onestep"` | estimated probabilities used as plug-ins (point estimates do not use the audit's true categories); standard errors include their uncertainty, for every design below |

The examples below use `"sub"`; the last sections compare it with the
other estimators.

## The two basic designs

| | Internal validation | External validation |
|---|---|---|
| Who is audited | a subsample of the rows used in the regression | units outside the regression data |
| What is recorded | true $Z$ for the audited rows | true $Z$ and proxy $\hat Z$ |
| Specification | `validation_sample(z, index = rows)` | `validation_sample(z, z_hat = proxies)` |
| What must carry over | nothing; the audit is part of the data | $\Pi$ (always) and $\pi$ (unless `prevalence = "em"`) |

**Categories are matched by label.** `z` and `z_hat` can be factors,
character vectors or numbers, and they are matched to the categories of
the `mc()` variable by their labels, never by position. If the proxy in
the data is a factor `occ` with levels `"clerk"`, `"manager"`, `"sales"`,
pass the audit as those labels, in any order:

```{r coding, eval = FALSE}
validation_sample(z = audit$true_occ, index = audit$row)   # "clerk", "sales", ...
```

Numbers are compared as labels too. They fit an integer-coded proxy
(categories `0, 1, ...`) or a factor whose levels are those numbers (e.g.
ISCO major groups `1`, ..., `9`), and are rejected with an explanatory
error for a proxy with other labels. A label that is not a category of the
model, such as a typo, is an error. The estimated $\Pi$, $\pi$ and $W$ are
labelled by category.

**Rows.** For internal validation, `index` gives row numbers of the data
passed to `mcglm()`. Subset or remove missing values **before** building
the index.

## A running example

```{r generator}
Pi_true <- matrix(c(0.88, 0.12,     # column 1: P(z_hat | Z = 0)
                    0.15, 0.85),    # column 2: P(z_hat | Z = 1)
                  2, 2)
simulate <- function(n, prev = 0.35, Pi = Pi_true) {
  z     <- rbinom(n, 1, prev)
  z_hat <- ifelse(z == 1, rbinom(n, 1, Pi[2, 2]), rbinom(n, 1, Pi[2, 1]))
  x1    <- rnorm(n)
  y     <- rpois(n, exp(-0.3 + 0.8 * z + 0.5 * x1))
  data.frame(y, z_true = z, z = z_hat, x1)
}
df <- simulate(5000)
```

The regression data `df` contain the proxy `z`; the true category `z_true`
is used below only to create audits and for comparison. The target is
$\gamma = 0.8$ with intercept $-0.3$.

```{r naive}
fit_naive <- mcglm(y ~ mc(z) + x1, data = df, method = "naive")
coef(summary(fit_naive))$naive
```

## Scenario 1: external validation from the same population

A separate gold-standard sample of 400 units, drawn the same way as the
regression data, has both the true and the predicted category.

```{r ext-srs}
df_gold <- simulate(400)
val1 <- validation_sample(z = df_gold$z_true, z_hat = df_gold$z)
val1
fit1 <- mcglm(y ~ mc(z) + x1, data = df, method = c("naive", "sub"),
              validation = val1)
summary(fit1)
```

Note that `mc(z)` has no matrix: the probabilities come from the audit.
Supplying a matrix (or `pi_z`, `p01`, ...) together with a validation
sample is an error, so the two sources cannot be mixed by accident.

The estimated quantities are kept in the fit:

```{r ext-srs-nuis}
fit1$nuisance$sub$Pi
fit1$nuisance$sub$pi_z
```

## Scenario 2: external validation from a different population

Often the gold-standard data come from elsewhere, e.g. a labelled
benchmark with a different occupational mix. Then only the classifier's
error rates $\Pi$ can be assumed to carry over, not the prevalence. Use
`prevalence = "em"`: $\pi$ is re-estimated by maximum likelihood from the
proxy frequencies in `df` given $\hat\Pi$ (the EM procedure of Saerens,
Latinne and Decaestecker, 2002).

```{r ext-shift}
df_bench <- simulate(400, prev = 0.6)          # different prevalence
val2 <- validation_sample(df_bench$z_true, df_bench$z)

fit2_audit <- mcglm(y ~ mc(z) + x1, data = df, method = "sub",
                    validation = val2)
fit2_em    <- mcglm(y ~ mc(z) + x1, data = df, method = "sub",
                    validation = val2,
                    mc_control = control_mc(prevalence = "em"))
rbind(prevalence_from_audit = coef(fit2_audit),
      prevalence_from_df    = coef(fit2_em))
```

```{r ext-shift-numbers, include = FALSE}
f3 <- function(x) formatC(unname(x), format = "f", digits = 3)
pi_audit  <- fit2_audit$nuisance$sub$pi_z[2]
pi_em     <- fit2_em$nuisance$sub$pi_z[2]
int_audit <- coef(fit2_audit)["(Intercept)"]
int_em    <- coef(fit2_em)["(Intercept)"]
```

Taking the prevalence from the benchmark gives
$\hat\pi = `r f3(pi_audit)`$, while the prevalence in the regression
data is `r f3(mean(df$z_true))`. The correction is then miscalibrated,
most visibly in the intercept (`r f3(int_audit)` against the true
$-0.3$). With `prevalence = "em"` the prevalence is re-estimated from `df`
as `r f3(pi_em)` and the intercept becomes `r f3(int_em)`.
If $\hat\Pi^{-1}\hat p$ falls outside the simplex, EM returns the boundary
estimate with a warning instead of clamping; the standard errors are then
not available.

## Scenario 3: internal simple random subsample

A random 10% of the regression rows are sent for manual coding.

```{r int-srs}
idx3 <- sample(nrow(df), 500)
val3 <- validation_sample(z = df$z_true[idx3], index = idx3)
fit3 <- mcglm(y ~ mc(z) + x1, data = df, method = "sub", validation = val3)
coef(summary(fit3))$sub
fit3$nuisance$sub$beta_equation
```

The proxies of the audited rows are read from `df`. The audited rows also
enter the regression with their **true** category, replacing the
corrected score (Yi, Yan, Liao and Spiegelman, 2019, eq. 23;
`beta_equation = "yi"`). This uses the extra information and is valid
whenever selection into the audit depends only on $(\hat Z, x)$.

## Scenario 4: audit stratified by the predicted category

Audits are usually drawn per predicted class, e.g. 250 ads predicted as
each occupation, regardless of how common the class is. The audited units
are then **not** representative: rare predicted classes are oversampled.
Declare the stratification with `strata = "z_hat"`; the weights
$N_j/n_j$ are computed from the proxy counts in `df`.

```{r int-strata}
idx4 <- unlist(lapply(0:1, function(j) sample(which(df$z == j), 250)))
val4 <- validation_sample(df$z_true[idx4], index = idx4, strata = "z_hat")
val4_ignored <- validation_sample(df$z_true[idx4], index = idx4)

est4 <- estimate_mc(val4, z_hat = df$z)
est4_ignored <- estimate_mc(val4_ignored, z_hat = df$z)
list(true = Pi_true, stratified = round(est4$Pi, 3),
     ignored = round(est4_ignored$Pi, 3))
```

Ignoring the design distorts $\hat\Pi$ because column proportions are
computed over a sample with the wrong mix of predicted classes, and it
distorts the audit's share of each true class:

```{r int-strata-pi}
rbind(stratified = est4$pi, ignored = est4_ignored$pi,
      truth = prop.table(table(df$z_true)))
```

The predictive matrix $W$, however, is the within-stratum proportion of
each true class either way, because sampling within predicted classes
does not change $\Pr(Z \mid \hat Z)$:

```{r int-strata-W}
all.equal(est4$W, est4_ignored$W)
```

`"sub"` with the prevalence taken from the audit uses only $W$, so for it
declaring or ignoring the strata gives the same fit. The design matters as
soon as $\hat\Pi$ or $\hat\pi$ are used on their own, for instance when
the prevalence is re-estimated from `df` with `prevalence = "em"`:

```{r int-strata-fit}
fit4 <- mcglm(y ~ mc(z) + x1, data = df, method = "sub", validation = val4)
fit4_ignored <- mcglm(y ~ mc(z) + x1, data = df, method = "sub",
                      validation = val4_ignored)
em <- control_mc(prevalence = "em")
fit4_em <- mcglm(y ~ mc(z) + x1, data = df, method = "sub",
                 validation = val4, mc_control = em)
fit4_ignored_em <- mcglm(y ~ mc(z) + x1, data = df, method = "sub",
                         validation = val4_ignored, mc_control = em)
rbind(declared = coef(fit4), ignored = coef(fit4_ignored),
      declared_em = coef(fit4_em), ignored_em = coef(fit4_ignored_em))
```

Declaring the design is therefore always the safe choice.

## Scenario 5: audit with known, unequal inclusion probabilities

Suppose the audit is a subsample of a nonprobability sample (e.g. scraped
job ads) drawn with inclusion probabilities that depend on the ad, for
instance oversampling ads with a high response or a low classifier
confidence. Pass the design weights $d_i = 1/\text{inclusion probability}$:

```{r int-weights}
p_incl <- ifelse(df$y == 0, 0.04, 0.15)          # depends on the outcome
idx5   <- which(runif(nrow(df)) < p_incl)
val5 <- validation_sample(df$z_true[idx5], index = idx5,
                          weights = 1 / p_incl[idx5])
fit5 <- mcglm(y ~ mc(z) + x1, data = df, method = "sub", validation = val5)
coef(summary(fit5))$sub
fit5$nuisance$sub$beta_equation
```

With design weights, `beta_equation = "auto"` switches to the
design-weighted estimating equation

$$
\sum_{i} U_i(\psi) + \sum_{i \in V} d_i\{S_i(\psi) - U_i(\psi)\} = 0,
$$

where the first sum runs over all rows of `df`, $V$ is the audit, $U$ the
corrected score and $S$ the score at the true category. It is unbiased for
any known inclusion probabilities, including selection that depends on
the outcome. Both equations can be requested explicitly with
`control_mc(beta_equation = "yi")` or `"weighted"`.

### Hajek or Horvitz--Thompson

`estimator` chooses how weighted totals are turned into proportions:

```{r ht}
val5_ht <- validation_sample(df$z_true[idx5], index = idx5,
                             weights = 1 / p_incl[idx5], estimator = "ht")
est5    <- estimate_mc(val5, z_hat = df$z)
est5_ht <- estimate_mc(val5_ht, z_hat = df$z)
rbind(hajek = est5$pi, ht = est5_ht$pi)
all.equal(est5$Pi, est5_ht$Pi)
fit5_ht <- mcglm(y ~ mc(z) + x1, data = df, method = "sub",
                 validation = val5_ht)
```

The misclassification matrix is a ratio of two weighted totals and is the
same under both. The prevalence differs: Hajek divides by $\sum d_i$, HT by
the frame size $N$ (for an internal audit, the number of rows of `df`).
Hajek is the usual choice; HT is exactly design-unbiased for totals when
$N$ is known.

## Scenario 6: external audit stratified by predicted class

When a stratified audit was drawn from a separate frame (e.g. last
quarter's ads), give the frame's stratum sizes:

```{r ext-strata}
df_frame <- simulate(20000)
idx6 <- unlist(lapply(0:1, function(j) sample(which(df_frame$z == j), 200)))
val6 <- validation_sample(df_frame$z_true[idx6], df_frame$z[idx6],
                          strata = "z_hat",
                          N_strata = c(`0` = sum(df_frame$z == 0),
                                       `1` = sum(df_frame$z == 1)))
fit6 <- mcglm(y ~ mc(z) + x1, data = df, method = "sub", validation = val6)
coef(summary(fit6))$sub
```

User-defined strata work the same way (`strata = region`, with
`N_strata` named by stratum, or with explicit `weights`). For an external
sample, standard errors use a stratum-centred, with-replacement variance.

## Estimate once, inspect, then fit

`estimate_mc()` can be run on its own to look at the estimates before any
regression is fitted:

```{r inspect}
est3 <- estimate_mc(val3, z_hat = df$z)
summary(est3)
```

`diagnose_mc()` reports the smallest singular value and condition number
of $\hat\Pi$, the number of audited units per true class, empty cells, and
whether the prevalence is on the boundary. Problems produce a warning by
default; use `control_mc(on_ill = "error")` to stop instead. A nearly
uninformative classifier is flagged:

```{r diagnose}
weak <- validation_sample(z     = rep(0:1, each = 100),
                          z_hat = c(rep(0:1, c(51, 49)), rep(0:1, c(49, 51))))
diagnose_mc(estimate_mc(weak, control = control_mc(on_ill = "none")))
```

The estimate can then be passed in the formula:

```{r reuse}
fit3_est <- mcglm(y ~ mc(z, est3) + x1, data = df, method = "sub")
all.equal(coef(fit3_est), coef(fit3))
```

An estimate that used the proxies of `df` (internal audits, or
`prevalence = "em"`/`"inverse"`) belongs to that data set; `mcglm()`
refuses to apply it to other data. An external estimate with
`prevalence = "validation"` can be reused across data sets.

## Propagating or ignoring the estimation uncertainty

By default (`variance = "delta"`), the regression and misclassification
estimating equations are stacked, so standard errors include the
uncertainty from a finite audit. `variance = "conditional"` treats the
estimates as known, which shows how much the audit size costs:

```{r variance}
val_small <- validation_sample(df_gold$z_true[1:150], df_gold$z[1:150])
fit_small <- mcglm(y ~ mc(z) + x1, data = df, method = "sub",
                   validation = val_small)
fit_small_cond <- mcglm(y ~ mc(z) + x1, data = df, method = "sub",
                        validation = val_small,
                        mc_control = control_mc(variance = "conditional"))
rbind(delta = fit_small$se$sub, conditional = fit_small_cond$se$sub)
```

## Comparing the designs

The table collects the estimates of $\gamma$ (true value 0.8) and the
intercept (true value $-0.3$) with their standard errors for the fits
above. Two benchmarks frame them: the infeasible regression on the true
category, and `"sub"` with the true misclassification probabilities.

```{r compare}
fit_true  <- glm(y ~ z_true + x1, family = poisson, data = df)
fit_known <- mcglm(y ~ mc(z, Pi_true) + x1, data = df, method = "sub",
                   pi_z = 0.35)

estimates <- function(fit) {
  if (inherits(fit, "glm")) {
    cf <- coef(summary(fit))
    g  <- "z_true"
  } else {
    cf <- coef(summary(fit))[[fit$method[length(fit$method)]]]
    g  <- "gamma"
  }
  c(gamma = cf[g, 1], se_gamma = cf[g, 2],
    intercept = cf["(Intercept)", 1], se_intercept = cf["(Intercept)", 2])
}

fits <- list(
  "true Z (infeasible)"               = fit_true,
  "naive"                             = fit_naive,
  "known Pi and pi"                   = fit_known,
  "1 external, same population"       = fit1,
  "2 external, prevalence from audit" = fit2_audit,
  "2 external, prevalence by EM"      = fit2_em,
  "3 internal SRS"                    = fit3,
  "4 internal, strata declared"       = fit4,
  "4 internal, strata ignored"        = fit4_ignored,
  "4 strata declared, EM prevalence"  = fit4_em,
  "4 strata ignored, EM prevalence"   = fit4_ignored_em,
  "5 internal, weights (Hajek)"       = fit5,
  "5 internal, weights (HT)"          = fit5_ht,
  "6 external, strata"                = fit6,
  "small external audit"              = fit_small
)
n_audit <- c(NA, NA, NA, nrow(df_gold), nrow(df_bench), nrow(df_bench),
             length(idx3), rep(length(idx4), 4), length(idx5),
             length(idx5), length(idx6), 150)
comparison <- data.frame(n_audit, t(sapply(fits, estimates)))
knitr::kable(comparison, digits = 3)
```

```{r compare-numbers, include = FALSE}
cmp <- function(row, col) formatC(comparison[row, col], format = "f", digits = 3)
```

What the table shows for this draw:

- The naive estimate of $\gamma$ is attenuated
  (`r cmp("naive", "gamma")`).
- Even with the true probabilities, the correction costs precision: the
  standard error of $\hat\gamma$ rises from
  `r cmp("true Z (infeasible)", "se_gamma")` with the true category to
  `r cmp("known Pi and pi", "se_gamma")`.
- Estimating the probabilities from an audit adds to it, and the smaller
  the audit, the larger the addition:
  `r cmp("1 external, same population", "se_gamma")` with 400 external
  units against `r cmp("small external audit", "se_gamma")` with 150.
- The internal SRS audit of `r length(idx3)` rows also contributes their
  true category to the regression; its standard error is
  `r cmp("3 internal SRS", "se_gamma")`.
- A misspecified design is not flagged by the standard errors; it moves
  the point estimates. With the benchmark's prevalence the intercept is
  `r cmp("2 external, prevalence from audit", "intercept")` against
  `r cmp("2 external, prevalence by EM", "intercept")` with EM. Ignoring
  the stratification leaves the default fit unchanged
  (`r cmp("4 internal, strata ignored", "gamma")`), but with EM it moves
  $\hat\gamma$ from `r cmp("4 strata declared, EM prevalence", "gamma")`
  to `r cmp("4 strata ignored, EM prevalence", "gamma")`.
- The audit with outcome-dependent inclusion probabilities (scenario 5)
  has the largest standard error,
  `r cmp("5 internal, weights (Hajek)", "se_gamma")`: the
  design-weighted equation protects against selection on the outcome at
  the price of variable weights.
- Scenario 6 lies
  `r formatC((0.8 - comparison["6 external, strata", "gamma"]) / comparison["6 external, strata", "se_gamma"], format = "f", digits = 1)`
  standard errors from the true $\gamma$, which a single draw can produce.

A single draw is shown; in repeated samples the correctly specified
designs are unbiased and their standard errors match the sampling spread
(see the package tests).

## Choice of estimator

The same audit can be used with the expectation correction and the
induced likelihood. Both use the whole response distribution; the induced
likelihood also lets the regression data inform the misclassification
probabilities.

```{r estimators}
gamma_table <- function(fit)
  rbind(estimate = sapply(fit$coefficients, function(b) b[["gamma"]]),
        se       = sapply(fit$se, function(b) b[["gamma"]]))
fit3_all <- mcglm(y ~ mc(z) + x1, data = df, method = c("sub", "ec", "il"),
                  validation = val3)
gamma_table(fit3_all)
```

With an outcome-dependent audit (scenario 5), `"ec"` and `"sub"` rely on
the design weights, while `"il"` does not need them because the selection
depends only on observed variables:

```{r estimators-weights}
fit5_all <- mcglm(y ~ mc(z) + x1, data = df, method = c("sub", "ec", "il"),
                  validation = val5)
g5 <- gamma_table(fit5_all)
g5
```

The standard error of $\hat\gamma$ is `r formatC(g5["se", "sub"], format = "f", digits = 3)`
for `"sub"`, `r formatC(g5["se", "ec"], format = "f", digits = 3)` for `"ec"` and
`r formatC(g5["se", "il"], format = "f", digits = 3)` for `"il"`: the
induced likelihood needs no design weights, whose variability inflates
the other two, and it also uses the regression data to estimate the
misclassification probabilities. In
this draw the `"ec"` estimate lies
`r formatC((0.8 - g5["estimate", "ec"]) / g5["se", "ec"], format = "f", digits = 1)`
standard errors from the true value; in repeated samples of this design
all three are unbiased and their standard errors match the sampling
spread (see the package tests).

## When the true category depends on the covariates

`"sub"`, `"ec"` and `"il"` (and `"cs"`) assume that $Z$ is independent of
the covariates $x$: they use one prevalence $\pi$ for every row. When, for
example, occupations differ by region or firm size, that assumption fails.
The corrected score `"cs_akn"` needs only $\Pi$ and stays consistent:

```{r dependence}
df_dep <- df
df_dep$z_true <- rbinom(nrow(df), 1, plogis(-0.8 + 1.2 * df$x1))
df_dep$z <- ifelse(df_dep$z_true == 1, rbinom(nrow(df), 1, Pi_true[2, 2]),
                   rbinom(nrow(df), 1, Pi_true[2, 1]))
df_dep$y <- rpois(nrow(df), exp(-0.3 + 0.8 * df_dep$z_true + 0.5 * df$x1))
idx_dep <- sample(nrow(df), 500)
fit_dep <- mcglm(y ~ mc(z) + x1, data = df_dep,
                 method = c("sub", "il", "cs_akn"),
                 validation = validation_sample(df_dep$z_true[idx_dep],
                                                index = idx_dep))
fit_dep_true <- glm(y ~ z_true + x1, family = poisson, data = df_dep)

dep <- cbind(
  "true Z (infeasible)" = c(coef(fit_dep_true)[c("z_true", "x1")],
                            sqrt(diag(vcov(fit_dep_true)))[c("z_true", "x1")]),
  sapply(c("sub", "il", "cs_akn"), function(m)
    c(coef(fit_dep, method = m)[c("gamma", "x1")],
      fit_dep$se[[m]][c("gamma", "x1")])))
rownames(dep) <- c("gamma", "x1", "se_gamma", "se_x1")
round(dep, 3)
```

```{r dependence-numbers, include = FALSE}
d3 <- function(m) formatC(dep["x1", m] - dep["x1", "true Z (infeasible)"],
                          format = "f", digits = 3, flag = "+")
```

The fit on the true category is the reference for this draw (its slope of
`x1` is `r formatC(dep["x1", 1], format = "f", digits = 3)`; the true value
is 0.5). Against it, `"sub"` and `"il"` move the slope by `r d3("sub")` and
`r d3("il")`: with $Z$ correlated with `x1`, a single prevalence makes them
attribute part of the effect of $Z$ to `x1`, and this bias does not shrink
with the sample size. `"cs_akn"` differs by `r d3("cs_akn")`, that is
`r formatC(abs(dep["x1", "cs_akn"] - dep["x1", 1]) / dep["se_x1", "cs_akn"], format = "f", digits = 1)`
of its standard errors
(`r formatC(dep["se_x1", "cs_akn"], format = "f", digits = 3)`), which is
ordinary sampling variation; in repeated samples it is unbiased. Its standard errors are larger: it needs neither
the independence of $Z$ and $x$ nor a model for $\Pr(Z \mid x)$.

Because the audit is internal, `mcglm()` also tested whether the true
category is associated with the covariates in the audit and warned:

```{r dependence-test}
fit_dep$independence_test
```

The remedy for `"sub"`, `"ec"` and `"il"` is a covariate-dependent
prevalence, a multinomial logit for $\Pr(Z \mid x)$ fitted with
`nnet::multinom` (Yi et al., 2019, Section 4.3):

```{r dependence-model}
fit_dep_pm <- mcglm(y ~ mc(z) + x1, data = df_dep,
                    method = c("sub", "il"),
                    validation = validation_sample(df_dep$z_true[idx_dep],
                                                   index = idx_dep),
                    mc_control = control_mc(prevalence_model = ~ x1))
dep_pm <- sapply(c("sub", "il"), function(m)
  c(coef(fit_dep_pm, method = m)[c("gamma", "x1")],
    fit_dep_pm$se[[m]][c("gamma", "x1")]))
rownames(dep_pm) <- c("gamma", "x1", "se_gamma", "se_x1")
round(cbind(dep, model_sub = dep_pm[, "sub"], model_il = dep_pm[, "il"]), 3)
```

```{r dependence-model-numbers, include = FALSE}
d3pm <- function(m) formatC(dep_pm["x1", m] - dep["x1", "true Z (infeasible)"],
                            format = "f", digits = 3, flag = "+")
```

With the prevalence model the slopes of `"sub"` and `"il"` differ from the
true-category fit by `r d3pm("sub")` and `r d3pm("il")`, and they keep
smaller standard errors than `"cs_akn"`. For an external audit, give its
covariates with `validation_sample(z, z_hat, data = audit_df)`, or estimate
the model from the regression data's proxies with
`control_mc(prevalence = "em", prevalence_model = ~ x1)`.

## Summary

| Audit design | `validation_sample()` call | Notes |
|---|---|---|
| External, same population | `(z, z_hat)` | assumes $\Pi$ and $\pi$ carry over |
| External, different population | `(z, z_hat)` + `control_mc(prevalence = "em")` | only $\Pi$ carries over |
| Internal SRS | `(z, index = rows)` | audited rows use their true category |
| Internal, fixed $n$ per predicted class | `(z, index = rows, strata = "z_hat")` | weights $N_j/n_j$ from the data |
| Internal, unequal known probabilities | `(z, index = rows, weights = 1/p)` | design-weighted equation; Hajek or `estimator = "ht"` |
| External, stratified | `(z, z_hat, strata = ..., N_strata = ...)` | stratum-centred variance |
| True category depends on covariates | any of the above + `control_mc(prevalence_model = ~ x1 + ...)`, or `method = "cs_akn"` | external audits with `prevalence = "validation"` need `validation_sample(data = )` |

## References

Saerens, M., Latinne, P. and Decaestecker, C. (2002). Adjusting the outputs
of a classifier to new a priori probabilities: a simple procedure.
*Neural Computation*, 14(1), 21--41. doi:10.1162/089976602753284446.

Yi, G. Y., Yan, Y., Liao, X. and Spiegelman, D. (2019). Parametric
regression analysis with covariate misclassification in main
study/validation study designs. *The International Journal of
Biostatistics*, 15(1), 20170002. doi:10.1515/ijb-2017-0002.
